---
title: "Lab 04: Instrumental Variables"
subtitle: "PAF 512 - Advanced Regression"
author: "Lab Solutions"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>

```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---


## Instructions

There are **6 sets of questions** to be answered. You can earn up to **100 points + 10 bonus points**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each question
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---

## The Policy Problem

This lab is loosely based on research examining the relationship between public housing assistance and health outcomes (Fertig & Reingold, 2007).

The "healthy lifestyle bias" presents a challenge when estimating the effect of public housing on health: households who pursue housing subsidies may be more or less attentive to their health compared to households who do not pursue subsidies. If this is true, their health behaviors and outcomes would differ regardless of whether they lived in public housing.

We use an instrumental variable approach to address this omitted variable bias.

> **Research Question:** Does time spent in public housing improve health outcomes for low-income households?

> **Hypothesis:** Households who have received public housing assistance for longer periods will report better health outcomes.

---

## Data

Data were collected across multiple cities in the US. The study population includes only households eligible for public housing—i.e., households whose income is below 80 percent of their city's median family income.

### Variable Descriptions

| Variable Name | Description |
|---------------|-------------|
| `HealthStatus` | Health status index (scale from 1-20, higher = better health) |
| `PublicHousing` | Number of years spent in public housing |
| `Supply` | Number of available public housing units per 100 eligible households in the city |
| `ParentsHealthStatus` | Health status of the individual's parents (scale 1-20) |
| `WaitingTime` | Average waiting time before obtaining public housing assistance in the city (months) |
| `Stamp` | Dollar amount of food stamps (TANF) consumed each month |
| `Age` | Age of the individual |
| `Race` | Race (1 = White, 2 = Black, 3 = Hispanic, 4 = Other) |
| `Education` | Education level (1 = High School, 2 = Diploma, 3 = Bachelor, 4 = Master) |
| `MaritalStatus` | Marital status (1 = Single, 2 = Married, 3 = Widow, 4 = Divorced) |
| `HealthBehavior` | Healthy lifestyle index (OMITTED VARIABLE - included for teaching purposes only) |

### Load Packages

```{r}
# Install packages if needed
# install.packages("stargazer")
# install.packages("AER")
# install.packages("dplyr")
# install.packages("knitr")
# install.packages("corrplot")

library(stargazer)    # Clean regression tables
library(AER)          # ivreg() for instrumental variables
library(dplyr)        # Data manipulation
library(corrplot)     # Correlation visualization
```

### Load Data

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/IV_lab.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)

# Preview the data
head(data)
```

---

## Lab Questions

### Question 1: Data Exploration

#### Q1a: Produce a summary statistics table using stargazer. (5 points)

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results = 'asis' to properly render the table in the knitted HTML file.

stargazer(data, 
          type = "html",
          #type = "text",
          title = "Summary Statistics", 
          digits = 2)

```

#### Q1b: What are the mean and standard deviation of the dependent variable (HealthStatus)? (2.5 points)

**The mean of HealthStatus is $10.41$ and the standard deviation is $2.80$. This indicates that, on average, individuals in our sample report a health status of approximately $10.41$ on the 1-20 scale, with moderate variation around this mean.**

#### Q1c: What are the mean and standard deviation of the policy variable (PublicHousing)? (2.5 points)

**The mean of PublicHousing is $29.37$ years and the standard deviation is $8.80$ years. This indicates that, on average, individuals in our sample have spent approximately $29$ years in public housing, with considerable variation (some spending much less time and others much more).**

---

### Question 2: The Naive Model

#### Q2a: Run a naive model that predicts the effect of public housing assistance (PublicHousing) on health status (HealthStatus). Include Race, Education, Age, and MaritalStatus as control variables. Report the results using stargazer. (5 points)

> **Note:** This model does not account for our omitted variable (healthy lifestyle), so the coefficient on PublicHousing is likely biased.

```{r, results='asis'}
# Convert categorical variables to factors
data$Race <- as.factor(data$Race)
data$Education <- as.factor(data$Education)
data$MaritalStatus <- as.factor(data$MaritalStatus)

# Run naive model
naive_model <- lm(HealthStatus ~ PublicHousing + Race + Education + Age + MaritalStatus, 
                  data = data)

stargazer(naive_model,
          type = "html",
          #type = "text",
          title = "Naive Model: Effect of Public Housing on Health Status",
          dep.var.labels = "Health Status",
          covariate.labels = c("Public Housing (Years)", 
                               "Race: Black", "Race: Hispanic", "Race: Other",
                               "Education: Diploma", "Education: Bachelor", "Education: Master",
                               "Age",
                               "Marital: Married", "Marital: Widow", "Marital: Divorced"),
          omit.stat = c("f", "ser"),
          digits = 3)
```

#### Q2b: Report the coefficient on PublicHousing, including its value, direction, and statistical significance. According to this naive model, how much does health status change for each additional year spent in public housing? (5 points)

> **Note:** Refer to the variable descriptions to ensure accuracy in your interpretation

**The coefficient on PublicHousing is $0.304$, which is positive and statistically significant at the $p < 0.01$ level. According to this naive model, for each additional year spent in public housing, health status increases by approximately $0.304$ points on the 1-20 scale. This suggests a positive relationship between time in public housing and health outcomes. However, this estimate is likely biased due to the omitted variable (healthy lifestyle).**

---

### Question 3: Identifying Potential Instruments

The dataset contains four variables that could potentially serve as instrumental variables:

1. **Supply** - Number of available public housing units per 100 eligible households in the city
2. **ParentsHealthStatus** - Health status of the individual's parents
3. **WaitingTime** - Average waiting time (in months) before obtaining public housing assistance in the city
4. **Stamp** - Dollar amount of food stamps (TANF) consumed each month

#### Q3a: What three criteria must a valid instrumental variable satisfy? Define each criterion in your own words. (10 points)

> **Note:** For each criterion, explain what it means and why it matters for the IV approach to work.

**1. Relevance (Correlation with Policy Variable):** The instrument must be correlated with the policy variable (`PublicHousing`). This matters because the instrument needs to be able to predict variation in the policy variable. If there is no correlation, the instrument cannot help us isolate the effect of the policy. We can test this empirically by examining the correlation between Z and X1, and by checking the F-statistic in the first stage regression.

**2. Exogeneity (No Correlation with Omitted Variable):** The instrument must be uncorrelated with the omitted variable (`HealthBehavior`/healthy lifestyle). This matters because if the instrument is correlated with the omitted variable, it will transmit that same bias to our estimates - defeating the entire purpose of using an instrument. We can partially assess this by examining correlations, but ultimately this assumption relies on theoretical justification.

**3. Exclusion Restriction (No Direct Effect on Outcome):** The instrument must affect the outcome variable (`HealthStatus`) only through its effect on the policy variable (`PublicHousing`), not directly. This matters because if the instrument has a direct effect on the outcome, we cannot cleanly attribute the estimated effect to the policy variable. This criterion cannot be tested empirically and must be argued theoretically.

---

### Question 4: Evaluating Potential Instruments

Examine the correlation matrix provided below.

> **Note:** This teaching dataset includes the omitted variable (HealthBehavior) so you can evaluate the instruments. In real-world research, you would never observe the omitted variable—that's precisely why it's omitted! Here, we include it for pedagogical purposes so you can see which instruments satisfy the exogeneity criterion.

```{r, fig.width=10, fig.height=8}
# Create correlation matrix for key variables
cor_vars <- data[, c("HealthStatus", "PublicHousing", "HealthBehavior", 
                     "Supply", "ParentsHealthStatus", "WaitingTime", "Stamp")]
cor_matrix <- cor(cor_vars, use = "complete.obs")

# Calculate p-values for significance testing
p_mat <- cor.mtest(cor_vars, conf.level = 0.95)

# Visualize correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", 
         diag = FALSE,
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0,0,2,0))

# Add significance stars
pvals <- p_mat$p
stars <- ifelse(pvals < 0.001, "***",
         ifelse(pvals < 0.01, "**",
         ifelse(pvals < 0.05, "*", "")))

n <- ncol(cor_matrix)
for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    if (stars[i, j] != "") {
      text(j - 0.35, n - i + 1 + 0.35,
           stars[i, j],
           col = "red",
           cex = 0.9)
    }
  }
}
```

#### Q4a: Which of the three IV criteria from Q3a can you test empirically using a correlation matrix, and which must be argued theoretically? Explain your reasoning. (5 points)

**Empirically Testable:**

- **Relevance:** We can test this by examining the correlation between each potential instrument and the policy variable (PublicHousing). A statistically significant correlation indicates the instrument is relevant.

- **Exogeneity:** In this teaching dataset, we can identify whether the potential instrument and the omitted variable (HealthBehavior) are correlated - to pass, they must not be. In practice (the real world), we cannot test this because the omitted variable is unobserved - hence "omitted."

**Must Be Argued Theoretically:**

- **Exclusion Restriction:** This criterion cannot be tested empirically. Even if we observe a correlation between an instrument and the outcome, we cannot determine whether this correlation operates through the policy variable or directly. We must make a theoretical argument for why the instrument would have no direct effect on the outcome.

#### Q4b: Using the correlation matrix, evaluate each of the four potential instruments against the three criteria. For each variable, state which criteria it meets and which it fails, citing specific correlation values as evidence. (15 points)

> **Note:** Organize your response by instrument (Supply, ParentsHealthStatus, WaitingTime, Stamp), and for each, address all three criteria.

**1. Supply (Public housing units per 100 eligible households)**

- **Relevance:** The correlation between `Supply` and `PublicHousing` is $0.11$, which is weak, but significant. **PASSES** the relevance criterion (weak instrument).
- **Exogeneity:** The correlation between `Supply` and `HealthBehavior` is $0.01$, which is essentially zero. **PASSES** the exogeneity criterion.
- **Exclusion Restriction:** Theoretically, the supply of public housing in a city affects how long someone stays in public housing (through availability), but should not directly affect individual health status. **Potentially PASSES** by theory.
- **Overall Assessment:** `Supply` passes the three criteria, making it a **POTENTIALLY VALID instrument**.

**2. ParentsHealthStatus (Health status of parents)**

- **Relevance:** The correlation between `ParentsHealthStatus` and `PublicHousing` is $0.07$, which is very weak but significant. **PASSES** the relevance criterion.
- **Exogeneity:** The correlation between `ParentsHealthStatus` and `HealthBehavior` is $0.1$, which is weak but significant. **FAILS** the exogeneity criterion.
- **Exclusion Restriction:** Theoretically problematic—parents' health status may directly affect an individual's health through genetics or learned health behaviors, violating the exclusion restriction. **FAILS** by theory.
- **Overall Assessment:** `ParentsHealthStatus` fails both exogeneity and exclusion restriction, making it an **INVALID instrument**.

**3. WaitingTime (Average waiting time for public housing in the city)**

- **Relevance:** The correlation between `WaitingTime` and `PublicHousing` is $0.55$, which is strong and positive. **PASSES** the relevance criterion.
- **Exogeneity:** The correlation between `WaitingTime` and `HealthBehavior` is $-0.02$, which is essentially zero. **PASSES** the exogeneity criterion.
- **Exclusion Restriction:** The average waiting time in a city is determined by administrative factors (housing supply, demand, bureaucratic processes) and should not directly affect an individual's health status. It affects health only through its effect on access to public housing. **PASSES** by theory.
- **Overall Assessment:** `WaitingTime` meets all three criteria, making it a **VALID instrument**.

**4. Stamp (Food stamps/TANF amount)**

- **Relevance:** The correlation between `Stamp` and `PublicHousing` is $-0.01$, which is essentially zero and not statistically significant. **FAILS** the relevance criterion.
- **Exogeneity:** The correlation between `Stamp` and `HealthBehavior` is $0.03$, which is very weak and not statistically significant. **PASSES** the exogeneity criterion.
- **Exclusion Restriction:** Theoretically problematic—food stamp benefits could directly affect health status through improved nutrition, independent of public housing. **FAILS** by theory.
- **Overall Assessment:** `Stamp` fails both relevance and exclusion restriction, making it an **INVALID instrument**.

#### Q4c: Based on your analysis in Q4b, which variable(s) appear to be valid instruments? Justify your answer. (5 points)

**Based on the analysis above, `WaitingTime` and `Supply` are the only valid instruments. They are the only variables that are:**

1. **Significantly correlated with `PublicHousing`**
2. **Are uncorrelated with the omitted variable `HealthBehavior`** 
3. **Have plausible theoretical arguments for the exclusion restriction**

**The other two variables fail either the relevance criterion, the exclusion restriction, or both.**

---

### Question 5: First Stage Regression

#### Q5a: For each valid instrument identified in Q4c, run a first-stage regression predicting PublicHousing using the instrument and the control variables (Race, Education, Age, MaritalStatus). Report the results using stargazer. (10 points)

> **Note:** Run a first-stage regression for each instrument you identified as valid in Q4c.

```{r, results='asis'}
# First stage regression using PublicHousing
first_stage_wt <- lm(PublicHousing ~ WaitingTime + Race + Education + Age + MaritalStatus, 
                  data = data)

# First stage regression using Supply
first_stage_sup <- lm(PublicHousing ~ Supply + Race + Education + Age + MaritalStatus, 
                  data = data)

stargazer(first_stage_wt,first_stage_sup,
          #type = "text",
          type = "html",
          title = "First Stage Regression",
          dep.var.labels = c("Public Housing"),
          covariate.labels = c("Waiting Time (Months)", 
                               "Supply",
                               "Race: Black", "Race: Hispanic", "Race: Other",
                               "Education: Diploma", "Education: Bachelor", "Education: Master",
                               "Age",
                               "Marital: Married", "Marital: Widow", "Marital: Divorced"),
          omit.stat = c("ser"),
          digits = 3)
```

#### Q5b: Report the F-statistic from each first-stage model. What is the rule of thumb for avoiding the "weak instruments problem"? Do your instruments pass this threshold? (5 points)

**The rule of thumb is having an F-statistic greater than 10. The F-statistic for the first stage regression using `WaitingTime`is $42.27$, which is well above the threshold of $10$. The F-statistic for the first stage regression using `Supply` is $2.36$.**

#### Q5c: Based on the F-statistics and coefficient significance, which instrument would you select for the second stage? Justify your choice. (5 points)

**I would select `WaitingTime` as the instrument for the second stage for the following reasons:**

1. **Strong F-statistic:** The F-statistic of $42.27$ far exceeds the threshold of $10$, indicating a strong first stage.

2. **Significant coefficient:** The coefficient on `WaitingTime` in the first stage is $0.795$ and is statistically significant at the $p < 0.01$ level. This means that for each additional month of average waiting time in a city, individuals spend approximately $0.80$ more years in public housing.

3. **Theoretical validity:** `WaitingTime` satisfies the exclusion restriction—city-level administrative waiting times are determined by factors unrelated to individual health behaviors.

---

### Question 6: Second Stage Regression (2SLS)

#### Q6a: Using the instrument you selected in Q5c, run a two-stage least squares regression using the `ivreg()` function from the AER package. Report the results using stargazer, showing both the naive model and the IV model side-by-side. (10 points)

> **Note:** The `ivreg()` syntax is: `ivreg(Y ~ X1 + Controls | Z + Controls, data = d)`. Variables to the left of `|` are in the outcome equation; variables to the right are the instruments plus controls. Controls appear on BOTH sides.

```{r, results='asis'}
# IV regression using ivreg() from AER package
# Syntax: ivreg(outcome ~ policy + controls | instrument + controls)
iv_model <- ivreg(HealthStatus ~ PublicHousing + Race + Education + Age + MaritalStatus | 
                  WaitingTime + Race + Education + Age + MaritalStatus, 
                  data = data)

# Display naive model and IV model side-by-side
stargazer(naive_model, iv_model,
          type = "html",
          #type = "text",
          title = "Comparison: Naive OLS vs. Instrumental Variables",
          column.labels = c("Naive OLS", "IV (2SLS)"),
          dep.var.labels = "Health Status",
          covariate.labels = c("Public Housing (Years)", 
                               "Race: Black", "Race: Hispanic", "Race: Other",
                               "Education: Diploma", "Education: Bachelor", "Education: Master",
                               "Age",
                               "Marital: Married", "Marital: Widow", "Marital: Divorced"),
          omit.stat = c("ser"),
          digits = 3)
```

#### Q6b: Interpret the coefficient on PublicHousing from the IV model. What is its value, direction, and statistical significance? What does this coefficient represent in terms of the causal effect of public housing on health status? (5 points)

> **Note:** The `ivreg()` function handles both stages internally. The coefficient on PublicHousing in the IV model represents the "cleaned" effect - using only variation in public housing that comes from your selected instrument, which is unrelated to the omitted variable (healthy lifestyle).

**The coefficient on `PublicHousing` in the IV model is $0.203$ and is statistically significant at the $p < 0.01$ level. This coefficient is positive, indicating that additional years in public housing are associated with better health outcomes.**

**Interpretation:** After using the instrumental variable to remove the bias caused by the omitted variable (healthy lifestyle), we find that each additional year spent in public housing increases health status by approximately $0.203$ points on the 1-20 scale. This represents the causal effect of public housing on health, isolated from the confounding influence of individuals' healthy lifestyle choices.

#### Q6c: According to your IV model, how much does health status change for each additional year spent in public housing? (5 points)

**According to the IV model, health status increases by $0.203$ points for each additional year spent in public housing. This means that:**

- **10 additional years** in public housing would increase health status by approximately $2.03$ points
- **20 additional years** in public housing would increase health status by approximately $4.06$ points

**On a 1-20 scale, these are meaningful improvements in health status.**

#### Q6d: Compare the coefficient on PublicHousing from your naive model (Q2) to the coefficient on PublicHousing from your IV model. Was the naive model over-estimating or under-estimating the effect of public housing? Based on the correlation structure (the relationships between PublicHousing, HealthBehavior, and HealthStatus), explain why the bias went in this direction. (5 points)

```{r}
# Compare coefficients
naive_coef <- coef(naive_model)["PublicHousing"]
iv_coef <- coef(iv_model)["PublicHousing"]

cat("Naive Model Coefficient:", round(naive_coef, 3), "\n",
    "IV Model Coefficient:", round(iv_coef, 3), "\n",
    "Difference:", round(naive_coef - iv_coef, 3), "\n",
    "Bias Direction:", ifelse(naive_coef > iv_coef, "OVERESTIMATED", "UNDERESTIMATED"))
```

**Comparison:**

- **Naive model coefficient:** $0.304$
- **IV model coefficient:** $0.203$
- **Difference:** $0.101$

**The naive model was OVERESTIMATING the effect of public housing on health status by approximately $0.101$ points per year.**

**Explanation of Bias Direction:**

The direction of the bias can be understood by examining the correlation structure:

1. **`PublicHousing` and `HealthBehavior` correlation:** Looking at the correlation matrix, `PublicHousing` is positively correlated with `HealthBehavior` (r = $0.811$). This means that people who spend more time in public housing also tend to have healthier lifestyles.

2. **`HealthBehavior` and `HealthStatus` correlation:** `HealthBehavior` is strongly positively correlated with `HealthStatus` (r = $0.904$). People with healthier lifestyles have better health outcomes.

3. **The bias mechanism:** In the naive model, when we don't control for `HealthBehavior`, some of the positive effect that healthy lifestyle has on health status gets incorrectly attributed to public housing. The naive model captures both:
   - The true effect of public housing on health
   - PLUS the spurious correlation caused by the omitted variable

4. **Result:** Because both correlations are positive (`PublicHousing`< - >`HealthBehavior` and `HealthBehavior` < - > `HealthStatus`), the omitted variable bias is **positive**, causing us to **overestimate** the true effect of public housing.

**The IV approach removes this bias by using only the variation in public housing that comes from `WaitingTime` - which is unrelated to healthy lifestyle choices - giving us a cleaner estimate of the true causal effect.**

---

### BONUS Question: Real-World Interpretation (10 points)

Imagine you are explaining your results to a city council considering expanding public housing.

#### BQ1a: From your summary statistics, identify the average years spent in public housing and the value at the 75th percentile.

```{r}
avg_housing <- mean(data$PublicHousing)
q75_housing <- quantile(data$PublicHousing, 0.75)

cat("Average years in public housing:", round(avg_housing, 1), "\n",
    "75th percentile:", round(q75_housing, 1), "\n",
    "Difference:", round(q75_housing - avg_housing, 1), "years")
```

**The average years spent in public housing is $29.4$ years, and the 75th percentile is $35.2$ years. The difference between these two values is approximately $5.9$ years.**


#### BQ1b: Using your IV model results, calculate the predicted difference in health status between someone at the average years of public housing and someone at the 75th percentile (holding all other variables constant). What does this difference mean in practical terms?

```{r}
# Calculate predicted health difference
years_difference <- q75_housing - avg_housing
health_difference <- iv_coef * years_difference

cat("Years difference (75th percentile - mean):", round(years_difference, 1), "\n",
    "IV coefficient:", round(iv_coef, 3), "\n",
    "Predicted health status difference:", round(health_difference, 2), "points")
```

**Calculation:**

- Years difference: $35.2 - 29.4 = 5.9$ years
- IV coefficient: $0.203$ points per year
- Health status difference: $5.9 \times 0.203 = 1.19$ points

**Practical Interpretation:**

Moving from average public housing tenure to the 75th percentile (an additional $5.9$ years) is associated with approximately a $1.2$-point increase in health status on the 1-20 scale.

**For the city council:** This suggests that public housing has a modest but real positive effect on health outcomes. While the effect per year is relatively small ($0.203$ points), it accumulates over time. Expanding public housing access could contribute to improved population health, though the health benefits alone may not justify the investment -other benefits of stable housing (economic stability, child educational outcomes, etc.) should also be considered.


---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-##-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Platforms like BlackBoard and Canvas sometimes disallow you from submitting HTML files when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```