---
title: "Lab 04: Instrumental Variables"
subtitle: "PAF 512 - Advanced Regression"
author: "Insert Your First and Last Name Here"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>

```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---


## Instructions

There are **6 sets of questions** to be answered. You can earn up to **100 points + 10 bonus points**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each question
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---

## The Policy Problem

This lab is loosely based on research examining the relationship between public housing assistance and health outcomes (Fertig & Reingold, 2007).

The "healthy lifestyle bias" presents a challenge when estimating the effect of public housing on health: households who pursue housing subsidies may be more or less attentive to their health compared to households who do not pursue subsidies. If this is true, their health behaviors and outcomes would differ regardless of whether they lived in public housing.

We use an instrumental variable approach to address this omitted variable bias.

> **Research Question:** Does time spent in public housing improve health outcomes for low-income households?

> **Hypothesis:** Households who have received public housing assistance for longer periods will report better health outcomes.

---

## Data

Data were collected across multiple cities in the US. The study population includes only households eligible for public housing—i.e., households whose income is below 80 percent of their city's median family income.

### Variable Descriptions

| Variable Name | Description |
|---------------|-------------|
| `HealthStatus` | Health status index (scale from 1-20, higher = better health) |
| `PublicHousing` | Number of years spent in public housing |
| `Supply` | Number of available public housing units per 100 eligible households in the city |
| `ParentsHealthStatus` | Health status of the individual's parents (scale 1-20) |
| `WaitingTime` | Average waiting time before obtaining public housing assistance in the city (months) |
| `Stamp` | Dollar amount of food stamps (TANF) consumed each month |
| `Age` | Age of the individual |
| `Race` | Race (1 = White, 2 = Black, 3 = Hispanic, 4 = Other) |
| `Education` | Education level (1 = High School, 2 = Diploma, 3 = Bachelor, 4 = Master) |
| `MaritalStatus` | Marital status (1 = Single, 2 = Married, 3 = Widow, 4 = Divorced) |
| `HealthBehavior` | Healthy lifestyle index (OMITTED VARIABLE - included for teaching purposes only) |

### Load Packages

```{r}
# Install packages if needed
# install.packages("stargazer")
# install.packages("AER")
# install.packages("dplyr")
# install.packages("knitr")
# install.packages("corrplot")

library(stargazer)    # Clean regression tables
library(AER)          # ivreg() for instrumental variables
library(dplyr)        # Data manipulation
library(knitr)        # Tables
library(corrplot)     # Correlation visualization
```

### Load Data

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/IV_lab.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)

# Preview the data
head(data)
```

---

## Lab Questions

### Question 1: Data Exploration

#### Q1a: Produce a summary statistics table using stargazer. (5 points)

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results = 'asis' to properly render the table in the knitted HTML file.

# Your Code Here
```

#### Q1b: What are the mean and standard deviation of the dependent variable (HealthStatus)? (2.5 points)

**Type your answer here, within the asterisks.**

#### Q1c: What are the mean and standard deviation of the policy variable (PublicHousing)? (2.5 points)

**Type your answer here, within the asterisks.**

---

### Question 2: The Naive Model

#### Q2a: Run a naive model that predicts the effect of public housing assistance (PublicHousing) on health status (HealthStatus). Include Race, Education, Age, and MaritalStatus as control variables. Report the results using stargazer. (5 points)

> **Note:** This model does not account for our omitted variable (healthy lifestyle), so the coefficient on PublicHousing is likely biased.

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results = 'asis' to properly render the table in the knitted HTML file.

# Your Code Here
```

#### Q2b: Report the coefficient on PublicHousing, including its value, direction, and statistical significance. According to this naive model, how much does health status change for each additional year spent in public housing? (5 points)

> **Note:** Refer to the variable descriptions to ensure accuracy in your interpretation

**Type your answer here, within the asterisks.**

---

### Question 3: Identifying Potential Instruments

The dataset contains four variables that could potentially serve as instrumental variables:

1. **Supply** - Number of available public housing units per 100 eligible households in the city
2. **ParentsHealthStatus** - Health status of the individual's parents
3. **WaitingTime** - Average waiting time (in months) before obtaining public housing assistance in the city
4. **Stamp** - Dollar amount of food stamps (TANF) consumed each month

#### Q3a: What three criteria must a valid instrumental variable satisfy? Define each criterion in your own words. (10 points)

> **Note:** For each criterion, explain what it means and why it matters for the IV approach to work.

**Type your answer here, within the asterisks.**

---

### Question 4: Evaluating Potential Instruments

Examine the correlation matrix provided below.

> **Note:** This teaching dataset includes the omitted variable (HealthBehavior) so you can evaluate the instruments. In real-world research, you would never observe the omitted variable—that's precisely why it's omitted! Here, we include it for pedagogical purposes so you can see which instruments satisfy the exogeneity criterion.

```{r, fig.width=10, fig.height=8}
# Create correlation matrix for key variables
cor_vars <- data[, c("HealthStatus", "PublicHousing", "HealthBehavior", 
                     "Supply", "ParentsHealthStatus", "WaitingTime", "Stamp")]
cor_matrix <- cor(cor_vars, use = "complete.obs")

# Calculate p-values for significance testing
p_mat <- cor.mtest(cor_vars, conf.level = 0.95)

# Visualize correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", 
         diag = FALSE,
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0,0,2,0))

# Add significance stars
pvals <- p_mat$p
stars <- ifelse(pvals < 0.001, "***",
         ifelse(pvals < 0.01, "**",
         ifelse(pvals < 0.05, "*", "")))

n <- ncol(cor_matrix)
for (i in 1:(n - 1)) {
  for (j in (i + 1):n) {
    if (stars[i, j] != "") {
      text(j - 0.35, n - i + 1 + 0.35,
           stars[i, j],
           col = "red",
           cex = 0.9)
    }
  }
}
```

#### Q4a: Which of the three IV criteria from Q3a can you test empirically using a correlation matrix, and which must be argued theoretically? Explain your reasoning. (5 points)

**Type your answer here, within the asterisks.**

#### Q4b: Using the correlation matrix, evaluate each of the four potential instruments against the three criteria. For each variable, state which criteria it meets and which it fails, citing specific correlation values as evidence. (15 points)

> **Note:** Organize your response by instrument (Supply, ParentsHealthStatus, WaitingTime, Stamp), and for each, address all three criteria.

**Type your answer here, within the asterisks.**

#### Q4c: Based on your analysis in Q4b, which variable(s) appear to be valid instruments? Justify your answer. (5 points)

**Type your answer here, within the asterisks.**

---

### Question 5: First Stage Regression

#### Q5a: For each valid instrument identified in Q4c, run a first-stage regression predicting PublicHousing using the instrument and the control variables (Race, Education, Age, MaritalStatus). Report the results using stargazer. (10 points)

> **Note:** Run a first-stage regression for each instrument you identified as valid in Q4c.

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results = 'asis' to properly render the table in the knitted HTML file.

# Your Code Here
```

#### Q5b: Report the F-statistic from each first-stage model. What is the rule of thumb for avoiding the "weak instruments problem"? Do your instruments pass this threshold? (5 points)

**Type your answer here, within the asterisks.**

#### Q5c: Based on the F-statistics and coefficient significance, which instrument would you select for the second stage? Justify your choice. (5 points)

**Type your answer here, within the asterisks.**

---

### Question 6: Second Stage Regression (2SLS)

#### Q6a: Using the instrument you selected in Q5c, run a two-stage least squares regression using the `ivreg()` function from the AER package. Report the results using stargazer, showing both the naive model and the IV model side-by-side. (10 points)

> **Note:** The `ivreg()` syntax is: `ivreg(Y ~ X1 + Controls | Z + Controls, data = d)`. Variables to the left of `|` are in the outcome equation; variables to the right are the instruments plus controls. Controls appear on BOTH sides.

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results = 'asis' to properly render the table in the knitted HTML file.

# Your Code Here
```

#### Q6b: Interpret the coefficient on PublicHousing from the IV model. What is its value, direction, and statistical significance? What does this coefficient represent in terms of the causal effect of public housing on health status? (5 points)

> **Note:** The `ivreg()` function handles both stages internally. The coefficient on PublicHousing in the IV model represents the "cleaned" effect - using only variation in public housing that comes from your selected instrument, which is unrelated to the omitted variable (healthy lifestyle).

**Type your answer here, within the asterisks.**

#### Q6c: According to your IV model, how much does health status change for each additional year spent in public housing? (5 points)
```{r}
# Your Code Here
# Use of code here is optional, but some math is required
```

**Type your answer here, within the asterisks.**

#### Q6d: Compare the coefficient on PublicHousing from your naive model (Q2) to the coefficient on PublicHousing from your IV model. Was the naive model over-estimating or under-estimating the effect of public housing? Based on the correlation structure (the relationships between PublicHousing, HealthBehavior, and HealthStatus), explain why the bias went in this direction. (5 points)

```{r}
# Your Code Here
# Use of code here is optional, but some math will be necessary
```

**Type your answer here, within the asterisks.**

---

### BONUS Question: Real-World Interpretation (10 points)

Imagine you are explaining your results to a city council considering expanding public housing.

#### BQ1a: From your summary statistics, identify the average years spent in public housing and the value at the 75th percentile.

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks.**


#### BQ1b: Using your IV model results, calculate the predicted difference in health status between someone at the average years of public housing and someone at the 75th percentile (holding all other variables constant). What does this difference mean in practical terms?

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks.**


---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-##-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Platforms like BlackBoard and Canvas sometimes disallow you from submitting HTML files when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```