---
title: "Lab 06: Logistic Regression"
subtitle: "PAF 512 - Advanced Regression"
author: "Insert Your First and Last Name Here"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>

```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **4 sets of questions** to be answered, plus a bonus question. You can earn up to **100 points + 10 bonus points**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each question
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---

## The Policy Problem

This lab is loosely based on the research conducted by Doan & Kirkpatrick (2013) entitled "Giving Girls a Shot: An Examination of Mandatory Vaccination Legislation" and published in *Policy Studies Journal*, 41(2), 295-318.

> **Research Question:** What factors influence a state's decision to introduce mandatory HPV vaccine legislation for young girls?

Doan and Kirkpatrick (2013) analyze state's decision to introduce a mandatory vaccine for the human papillomavirus (HPV) for all fifth- and sixth-grade girls in public schools. The vaccine was approved in 2006 and patented by Merck & Co. The two researchers were particularly interested in the influence of competing morality, economic, and public-health determinants on states' decision making.

With regards to morality, they suggest that state politicians are responsive to the public opinion within their own state, such that they will respond to positive or negative approval of the new legislation. For instance, a high proportion of Evangelics or Catholics within the state might discourage mandatory HPV vaccine legislation as those groups have conservative view regarding sexuality and sexual activity and might disapprove the vaccine at such early age. Similarly, a high rate of negative media attention might decrease politicians' motivation to introduce legislation mandating the HPV vaccine in order to remain aligned with the electorate's preferences.

By contrast, we could expect politicians to support their party's values. "Social welfare" is generally an issue associated with Democrats, which might have greater incentives to pass a mandatory HPV vaccine regulation to appeal their electoral basis as compared to Republicans.

Finally, politicians might respond to political interests and support policies in favor of private organizations providing resources for their electoral campaign. We might expect that states where politicians receive more money from Merck & Co will be more likely to adopt such legislation.

> **Hypotheses:**
>
> - Media salience is negatively correlated with mandatory legislation.
> - Proportion of state legislature that is Democrat is positively correlated with mandatory legislation.
> - Religiosity (percentage of the population that is Catholic or Evangelical) is negatively correlated with mandatory legislation.
> - Merck PAC contributions are positively correlated with mandatory legislation.

Given that the dependent variable is a binary outcome (1/0), a probability model is appropriate. Answer the questions below using a linear probability model and a logistic regression model.

---

## Data

### Load Packages

```{r}
# Install packages if needed
# install.packages("stargazer")
# install.packages("sandwich")
# install.packages("lmtest")
# install.packages("margins")

library(stargazer)    # Clean regression tables
library(sandwich)     # Robust standard errors
library(lmtest)       # Coefficient testing with robust SEs
library(margins)      # Marginal effects for logit models
```

### Load Data

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/logit-lab.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)

head(data)
```

### Pre-Scale Proportion Variables in Decimal Form
> **Note: Important Data Scaling Issue**
>
> In this dataset, **Democrats**, **Evangelics**, and **Catholics** are recorded as decimals between 0 and 1 (0.50), and represent the proportion of the population in each respective group.
>
> **The Problem:** A "1-unit increase" in these variables means going from 0 to 1, which represents going from 0% to 100% of the population. That's unrealistic - we care about smaller changes like going from 30% to 31% so our results are more meaningful and interpretable.
>
> **The loaded data has been completed for you already, but the following two options are how you can fix it on your own in the future. **
>
> **How to Fix It:**
> - **Option One:** Convert each coefficient to be at the 1 percentage point increase level by dividing by 100. This requires many more steps throughout the modeling process and opens the door for additional errors. 
> - **Option Two:** Scale your entire column to no longer be constrained between 0 and 1, instead, 0 and 100 by multiplying by 100 at the column variable level. Once done, our data contains pre-scaled variables for a 1-percentage point unit change. 
>
>
> **Bottom Line:** Your data contains columns with scaled variables (`Democrats_s`, `Evangelics_s`, `Catholics_s`), be sure to use those.

```{r}
# Create new columns with our scaled variables
## We only scale the variables that are proportions and in decimal form
data$Democrats_s <- round(data$Democrats * 100, 5)
data$Evangelics_s <- round(data$Evangelics * 100, 5)
data$Catholics_s <- round(data$Catholics * 100, 5)

# Preview Results
head(data)
```

```{r}
# Confirm scaled variables are contained between 0 and 100 now
summary(data)
```


### Variable Descriptions

| Variable Name | Description |
|---------------|-------------|
| `Adoption` | Whether a state legislature has considered a mandatory vaccine law (=1) or not (=0) |

| `Democrats` | Percentage of House Democrats in the state legislature (0-1) |
| `Democrats_s` | Scaled percentage of House Democrats in the state legislature (0-100) |

| `Evangelics` | Percentage of Evangelical population (0-1) |
| `Evangelics_s` | Scaled percentage of Evangelical population (0-100) |

| `Catholics` | Percentage of Catholic population (0-1) |
| `Catholics_s` | Percentage of Catholic population (0-100) |

| `Media` | Number of articles covering the HPV vaccine issue in major news sources in a state in the past year |
| `Merck` | Average total dollar amount of contributions given to candidates for state offices |


#### Null non-scaled variables to prevent confusion
```{r}
# Prevent the use of these columns accidentally
data$Democrats <- NULL
data$Evangelics <- NULL
data$Catholics <- NULL
```


---

## Lab Questions

### Question 1: The Linear Probability Model (30 points)

#### Q1a: Run a linear probability model (LPM) predicting Adoption using all five independent variables. Present the results using stargazer. (5 points)

> **Note:** A linear probability model is simply an OLS model (`lm()`) where the dependent variable is binary (0/1).

```{r, results='asis'}
# Reminder: While completing your lab, set #type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Your Code Here
```

#### Q1b: Calculate robust standard errors for the LPM. Display both the non-robust and robust models side-by-side using stargazer. (5 points)

> **Note:** The LPM suffers from heteroskedasticity by construction. Robust standard errors correct for this. Use `vcovHC()` from the `sandwich` package and supply the robust SEs to stargazer using the `se = list()` argument.

```{r, results='asis'}
# Reminder: While completing your lab, set #type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Your Code Here
```

#### Q1c: Which variables are statistically significant predictors of adoption? For each significant variable, report the coefficient name, its raw value, the estimated change in probability in percentage points (e.g., 0.05 becomes 5 percentage points), significance level, and the direction of its effect on the probability of adoption. (10 points)

> **Note:** Remember that Democrats_s, Evangelics_s, and Catholics_s were pre-scaled.

```{r}

# Show your work, if desired, when converting to % change

```

**Type your answer here, within the asterisks**

#### Q1d: Robust standard errors solve the heteroskedasticity problem in LPMs. What are the other two limitations of the linear probability model? (10 points)

**Type your answer here, within the asterisks**

---

### Question 2: The Logit Model (20 points)

#### Q2a: What estimator does the logit model use instead of OLS, and what shape does the predicted probability curve take? Briefly explain why this shape is preferred over the linear probability model. (5 points)

**Type your answer here, within the asterisks**

#### Q2b: Run the logit model using `glm()` with `family = "binomial"` and present BOTH the LPM and logit models side-by-side using stargazer. (10 points)
> **Note:** Remember that Democrats_s, Evangelics_s, and Catholics_s were pre-scaled.

```{r, results='asis'}
# Reminder: While completing your lab, set #type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Your Code Here
```

#### Q2c: Can you directly compare the coefficient values between the LPM and logit models? Why or why not? (5 points)

**Type your answer here, within the asterisks**

---

### Question 3: Interpreting Logit Coefficients (30 points)

#### Q3a: Transform the logit model coefficients into odds ratios using `exp()`. Report and interpret the odds ratio for the Evangelics variable. (5 points)

> **Note:** Remember that Democrats_s, Evangelics_s, and Catholics_s were pre-scaled.

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**


#### Q3b: Use the probability formula to calculate the predicted probability of adopting mandatory vaccine legislation for an average state (i.e., a state where all variables are set at their means). Show your work. (10 points)
> **Note:** Plug in the raw mean of each independent variable and the coefficients from your logit model. Do not use `predict()` for this question.

$$
P = \frac{1}{1 + e^{-(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3)}} \tag{3.1}
$$

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**


#### Q3c: Compared to the average state, how much does the probability change for a state that has a particularly high media coverage, equal to the 95th percentile? Use the `predict()` function. Calculate the predicted probability for the average state, the percentile state, and the difference (discrete change) between the two. (5 points)
> **Note:** make sure you column names match the '_s' convention. Predict looks for the same named variables in your logit model, if they're different you will get an error.

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

#### Q3d: Compared to the average state, how much does the probability change for a state with a particularly low percentage of Evangelicals, equal to the 25th percentile? Use the `predict()` function. Calculate the predicted probability for the average state, the percentile state, and the difference (discrete change) between the two. (5 points)
> **Note:** Remember to name your new data frame columns the same as your logit model variable names or you will get an error.

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

#### Q3e: Create a plot showing how the predicted probability of adoption changes as the percentage of Evangelicals ranges from 0 to 100. Hold all other variables at their means. (5 points)

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

---

### Question 4: Marginal Effects (20 points)

#### Q4a: Calculate the average marginal effects (AME) for all variables in the logit model using the `margins()` function. Which variable has the highest average marginal effect between Media and Evangelic? Report both AME values. (5 points)
> Note: We're looking for the largest **absolute** value

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

#### Q4b: Does an increase in the Evangelical population have a greater effect in a state where the current percentage is 20% or 40%? Calculate the marginal effect of Evangelic at both values and explain why they differ. (10 points)
> **Note:** Remember to use our pre-scaled variable. Use rbind() to vertically stack both objects for a nicer display.

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

#### Q4c: Imagine you work for an advocacy group trying to **promote the adoption** of mandatory HPV vaccine legislation across states. Part of your effort is to decrease negative media coverage. Would you rather invest in a state where the media coverage is equal to 200 or in a state where the media coverage is equal to 400? Calculate the marginal effect at both values and explain your recommendation. (5 points)

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

---

### BONUS: LPM vs. Logit Out-of-Bound Predictions (10 points)

#### BQ1a: Using your LPM model from Q1, calculate the predicted probability of adoption for a state with the following profile: Democrats = 90, Evangelic = 5, Catholics = 5, Media = 50, Merck = 5000. Now calculate the predicted probability using your logit model from Q2 for the same state. What do you notice about the LPM prediction? Why is this problematic, and how does the logit model solve it? (10 points)

```{r}
# Your Code Here
```


**Type your answer here, within the asterisks**

---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-06-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Platforms like BlackBoard and Canvas sometimes disallow you from submitting HTML files when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```
