---
title: "Lab 05: Regression Discontinuity"
subtitle: "PAF 512 - Advanced Regression"
author: "Insert Your First and Last Name Here"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>

```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **5 sets of questions** to be answered, plus bonus questions. You can earn up to **100 points + 10 bonus points**. Points are indicated next to each question.

Remember to:

- With the exception of `rdrobust()` models, format your tables using `stargazer`
- Submit the R markdown file & HTML with all your codes
- **Bold your answers**

---

## The Policy Problem

**Research Question:** Does summer school improve low-achieving students' grades?

Summer school programs are designed to help students who have not yet achieved the skills required for the next level. For this lab, we are interested in testing the following:

**Hypothesis:** Students who attend summer school will achieve higher grades the following year.

Students that scored below 60 on the 8th grade standardized math exam were required to attend summer school before 9th grade. Students scoring 60 or above were not required to attend.

> **Note:** As we learned in the codethrough, checking whether the cutoff was adhered to is an important first step in delineating between Fuzzy vs. Sharp design. In the case of this dataset, the design was sharp - no students that scored above 59 attended summer school and all students that scored below 60 did.

---

## Data

The dataset contains observations on 1,000 high school students. Variables include:

| Variable | Description |
|----------|-------------|
| `math_7` | Score on standardized math exam at end of 7th grade (0-100) |
| `gpa_8` | Math GPA during 8th grade, prior to summer school (0-4) |
| `math_8` | Score on standardized math exam at end of 8th grade (0-100) |
| `gpa_9` | Math GPA during 9th grade, following summer school (0-4) |

---

### Load Packages

```{r}
# Install packages if needed
# install.packages("stargazer")
# install.packages("rdrobust")

library(stargazer) # nice tables
library(rdrobust) # RDD Gold Standard
```


### Load Data

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/regression-discontinuity-lab.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)

# Preview the data
head(data)
```



---

## Lab Questions

### Question 1: Why Regression Discontinuity? (20 points)

Before running the model, we need to understand why a simple comparison of treatment and control groups would be misleading.

#### Q1a: Create a treatment variable called `treat` that equals 1 for students who scored below 60 on the 8th grade standardized math exam (and thus attended summer school), and 0 otherwise.(5 points)

```{r}
# Your Code Here
```

#### Q1b: Calculate the mean 9th grade math GPA for the treatment group and the control group separately. (5 points)

```{r}
# Your Code Here
```


#### Q1c: Conduct a t-test to determine whether the difference in mean 9th grade GPA between groups is statistically significant. Report the t-statistic and p-value. (5 points)

```{r}
# Your Code Here
```

**Type your answer here, within the asterisks**

#### Q1d: Based on Q1b and Q1c alone, it appears that summer school students have *lower* GPAs than non-summer school students. Explain why this naive comparison is misleading. What is the fundamental problem with comparing these two groups directly? (5 points)

> **Note:** Think about what determined which students attended summer school in the first place, and how that relates to the outcome we're measuring.

**Type your answer here, within the asterisks**

---

### Question 2: Preparing the Data (10 points)

#### Q2a: Write out the regression discontinuity model formula using proper notation. Define each term. (3 points)

> **Note:** Your formula should include: the outcome variable, a treatment indicator, and the rating variable centered at the cutoff.

**Type your answer here, within the asterisks**

#### Q2b: Create a new variable called `math8_c` that centers the 8th grade math score at the cutoff of 60. (3 points)

```{r}

# Your Code Here

```

#### Q2c: Why do we center the rating variable at the cutoff? What does the intercept represent in a model with centered versus non-centered rating variables? (4 points)

**Type your answer here, within the asterisks**

---

### Question 3: Estimating the Model (35 points)

#### Q3a: Estimate a regression discontinuity model using `lm()` that predicts 9th grade math GPA using your treatment variable and centered rating variable. Present your results using stargazer. Define and explain each coefficient. (8 points)
> **Note:** The data used in this lab is simulated. The slopes on both sides of the cutoff are identical. Therefore, for the sake of simplicity, we will use the Naive Method (parallel slopes) found in the lecture materials for this `lm()` model (i.e., do not use an interaction term). This assumes the relationship between math scores and GPA is identical for both groups, which simplifies our manual calculations later.

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), 
# but remember to set type = "html" when you are ready to knit and submit your lab.

# Your Code Here

```

**Type your answer here, within the asterisks**


#### Q3b: The basic `lm()` approach has limitations for RDD. It uses all observations regardless of distance from the cutoff, assumes a linear relationship across the entire range of scores, and may produce incorrect standard errors. Why might focusing only on observations *close to the cutoff* provide a more credible estimate of the treatment effect? (5 points)

**Type your answer here, within the asterisks**

#### Q3c: Now estimate the treatment effect using the `rdrobust()` function from the `rdrobust` package. Report the robust treatment effect estimate, the robust 95% confidence interval, and the bandwidth selected by the algorithm. (8 points)

> **Two Notes:** 
1) Unlike `lm()`, you do NOT need to center the rating variable or create a treatment dummy for `rdrobust()`. The function handles this internally. Use the original `math_8` variable and specify `c = 60`. Remember as well, `stargazer()` and `rdrobust()` do not get along - stick with `summary()` for your output.
2) Recall from the codethrough that rdrobust always calculates Right - Left. In our codethrough the treatment was on the right and control on the left. In this example, our treatment is on the left, while our control is on the right. Be mindful of your interpretation. Use `rdplot()` to view the change at the cutoff if you need more help.

```{r}

# Your Code Here

```


**Type your answer here, within the asterisks**

#### Q3d: Compare your results from the `lm()` model (Q3a) and the `rdrobust()` model (Q3c). Report the treatment effect coefficient name, its value, and statistical significance from each approach. Are they similar or different? Which estimate should we trust for making causal claims, and why? (8 points)

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**


#### Q3e: Create a regression discontinuity plot using the `rdplot()` function. Describe what you see. Is there a visible "jump" at the cutoff? Does the size of the jump appear consistent with your treatment effect estimate? (6 points)
> **Note:** You will likely receive a "Mass points detected in the running variable." note. Do not worry. Reference the codethrough for more details. 

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**

---

### Question 4: The Counterfactual (20 points)

The power of RDD is that it lets us estimate what *would have happened* to treated students if they had not received treatment.

> **Note:** We will use the coefficients from your `lm()` model in Q3a for these calculations. While `rdrobust()` gives a more accurate final estimate, the `lm()` model provides simple linear coefficients ($y = \beta_0 + \beta_1 x$) that make the logic of the counterfactual easier to see mathematically.

#### Q4a: According to your `lm()` model from Q3a, for each additional point a student scores on the 8th grade standardized math exam, how much does their predicted 9th grade math GPA change? Report the coefficient name, value, and interpretation. (5 points)

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**

#### Q4b: Consider a student who scored exactly 60 on the 8th grade math exam (right at the cutoff) and did NOT attend summer school. Using your `lm()` model coefficients, calculate their predicted 9th grade math GPA. Show your work. (5 points)

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**

#### Q4c: Now consider an identical student who scored exactly 60 but DID attend summer school. Calculate their predicted 9th grade math GPA. Show your work. (5 points)

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**

#### Q4d: The difference between your answers in Q4c and Q4b represents the treatment effect at the cutoff. Does this match your treatment coefficient from Q3a? Explain why these values should be the same. (5 points)

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**

---

### Question 5: Model Assumptions (15 points)

#### Q5a: The basic RDD model assumes a linear relationship between the rating variable and the outcome on both sides of the cutoff. Looking at your plot from Q3e, does this assumption appear reasonable for this data? What pattern would concern you? (7 points)

**Type your answer here, within the asterisks**

#### Q5b: A key assumption of RDD is that individuals cannot precisely manipulate their score to land on a preferred side of the cutoff. Generate a histogram of the running variable (math_8) to check for "bunching" around the cutoff. Does the density appear smooth? (8 points)

```{r}

# Your Code Here

```

**Type your answer here, within the asterisks**

---

### Bonus Question 1: Bandwidth Sensitivity Analysis (10 points)

In the codethrough, we learned that `rdrobust()` automatically selects an optimal bandwidth. However, a robust finding should not depend entirely on one specific bandwidth choice.

#### BQ1a: Run two additional `rdrobust()` models: one using half the optimal bandwidth and one using double the optimal bandwidth. Report the treatment effects and p-values for all three models (Half, Optimal, Double). (8 points)

> **Hint:** You can extract the optimal bandwidth from your model in Q3c using `rd_model$bws[1, 1]`. Use the argument `h = ...` to set a manual bandwidth. Use `coef[1]` for the treatment effect estimate.

```{r}

# Your Code Here

```

#### BQ1b: Does the treatment effect remain stable across these different bandwidths? Why is this check important for validation? (2 points)

**Type your answer here, within the asterisks**



---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-##-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Platforms like BlackBoard and Canvas sometimes disallow you from submitting HTML files when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```
