---
title: "Lab 02: Difference-In-Difference"
subtitle: "PAF 512 - Advanced Regression"
author: "Insert Your First and Last Name Here"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>
```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **4 sets of questions** to be answered. You can earn up to **100 points + bonus questions**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each questions
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---

## The Policy Problem

This lab is loosely based on Branas et al. (2011), entitled "A Difference-in-Differences Analysis of Health, Safety, and Greening Vacant Urban Space".

The 'broken windows' theory suggests that vacant lots offer refuge to criminal and other illegal activity and visibly symbolize that a neighborhood has deteriorated. To prevent these problems, **City A** has decided to implement a public program that will "green" some of the city's vacant lots in 4 different neighborhoods. 'Greening' includes removing trash and debris and planting grass and trees to create a park.

Three years after implementation, the city wants to know whether the program has been successful. They propose comparing lots that have been 'greened' (treatment) to lots that are still vacant but could have been chosen for greening (control).

> **Research Question:** Does greening vacant lots decrease vandalism in City A?

> **Hypothesis:** Greening vacant lots will decrease vandalism in City A.

---

## Data

Data on vandalism acts were reported by the local police and include the total number of calls about vandalism acts received during the year before and after the greening from households near the vacant lots.

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/diff-in-diff-lab.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)
head(data)
```

### Variable Descriptions

| Variable Name | Description |
|---------------|-------------|
| Vandalism | Number of calls about vandalism acts near the vacant lots |
| Group | Treatment (=1) and control group (=0) |
| Post | Observation before treatment (=0) and after treatment (=1) |

### Load Packages

```{r}
# Install/Load Packages
# install.packages("stargazer") 
# install.packages("dplyr")

library(stargazer)    # Clean regression tables
library(dplyr)        # Data manipulation
```

---

## Lab Questions

### Question 1: The DID Model

#### Q1: Write down the difference-in-difference formula and build the model using the variables in our dataset.
  > **Note: **Write the formula in the asterisks, build the model using lm() in the code chunk. Extra kudos if you utilize LaTeX for your formula as shown in the codethrough.
  
```{r}
# Your code here
```

**Type your answer here, within the asterisks**

---

### Question 2: Model Estimation and Interpretation

#### Q2a: Run the DID regression and present results using stargazer.

```{r}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include results='asis' in your code chunk header to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Additional Reminder: If you choose to use intercept.bottom = FALSE along with labeling your covariates manually, please be very cognizant of the order of your variables. This argument moves the Intercept to the TOP of the list, so your covariate.labels must start with the Intercept label to match.

# Your code here

```


#### Q2b: What is the average number of vandalism calls in the control group before the treatment? Which coefficient provides this information? (5 points)

**Type your answer here, within the asterisks**

#### Q2c: Did the treatment and control groups differ significantly in vandalism calls before the intervention? Which coefficient tells us this, and what is its value and significance? (10 points)

**Type your answer here, within the asterisks**

#### Q2d: Do treatment and control groups need to be identical in the pre-treatment period for valid DID inference? Explain why or why not. (10 points)

**Type your answer here, within the asterisks**

#### Q2e: Does the number of vandalism calls in the control group change after the intervention? Report the coefficient, whether it increases or decreases, and its significance. (5 points)

**Type your answer here, within the asterisks**

#### Q2f: What is the average number of vandalism calls in the control group after the intervention? Show the formula and calculation. (5 points)
  > Feel free to only type the answer, code is optional here.
  
```{r}
# Your code here
```


**Type your answer here, within the asterisks**

#### Q2g: Does greening help reduce vandalism? Report the coefficient name, value, significance, and interpret its real-world meaning. (10 points)

**Type your answer here, within the asterisks**

#### Q2h: What is the average number of vandalism calls in the treatment group after the intervention? Show the formula and calculation. (5 points)
  > Feel free to only type the answer, code is optional here.

```{r}
# Your code here
```

**Type your answer here, within the asterisks**

---

### Question 3: Counterfactual Analysis

#### Q3a: Write the counterfactual formula using beta coefficients. What does the counterfactual represent? (5 points)

**Type your answer here, within the asterisks**

#### Q3b: Calculate the counterfactual value. How many vandalism calls would we expect if the intervention had not occurred? (5 points)
  > Feel free to only type the answer, code is optional here.

```{r}
# Your code here
```

**Type your answer here, within the asterisks**

#### Q3c: What is the main assumption behind the counterfactual in a DID design? Why is this assumption important? (5 points)

**Type your answer here, within the asterisks**

---

### Question 4: Comparing DID and ITS Approaches

Now let's compare two different approaches to analyzing policy interventions using a different dataset with monthly observations.

```{r}
## Load and preview the data
URL2 <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/time-series-diff-n-diff-comparison.csv"
dat <- read.csv(URL2, stringsAsFactors = FALSE)
rbind(head(dat, 3), tail(dat, 3))
```

This dataset contains 48 months of data. The vacant lot clean-up occurred between months 24 and 25.

#### Visualize the data

```{r}

# Calculate averages by group and time
ts_control <- dat %>% 
  filter(group == "control") %>%
  group_by(time) %>% 
  summarize(ave = mean(crimes), .groups = "drop")

ts_treat <- dat %>% 
  filter(group == "treat") %>%
  group_by(time) %>% 
  summarize(ave = mean(crimes), .groups = "drop")

# Plot
plot(ts_control$time, ts_control$ave, type = "b", pch = 19, col = "steelblue",
     xlab = "Time (months)", ylab = "Number of Crimes",
     main = "Crime Rates Over Time by Group",
     xlim = c(0, 50), ylim = c(60, 225), bty = "n")
points(ts_treat$time, ts_treat$ave, type = "b", pch = 19, col = "firebrick")
abline(v = 24.5, lty = 3, col = "gray", lwd = 3)

legend("topleft", legend = c("Control", "Treatment"), 
       col = c("steelblue", "firebrick"), pch = 19, lty = 1)
```


```{r}
ts1 <- 
  dat %>% 
  filter( group == "control" ) %>%
  group_by( time ) %>% 
  summarize( ave=mean( crimes ) )

ts2 <- 
  dat %>% 
  filter( group == "treat" ) %>%
  group_by( time ) %>% 
  summarize( ave=mean( crimes ) )

y.24 <- ts2$ave[24]
y.25 <- ts2$ave[25]
diff <- y.24 - y.25 

plot( ts1$time, ts1$ave, type="b", pch=19, col="steelblue", bty="n",
      xlab="Time (months)", ylab="Number of Crimes", 
      xlim=c(0,50), ylim=c(60,225) )
points( ts2$time, ts2$ave, type="b", pch=19, col="firebrick" ) 
abline( v=c(12,36,48), lty=2, col="gray" )
abline( v=24.5, lty=3, col="gray", lwd=3 )
abline( h=c(y.24,y.25), lty=3, col="firebrick" )

arrows( x0=38, y0=y.24, y1=y.25, col="firebrick", code=3 )
text( x=40, y=mean(c(y.24,y.25)), 
      paste0( "Diff ~ ", round(diff,0) ), 
      col="firebrick", pos=4, cex=1.4 )
legend("topleft", legend = c("Control", "Treatment"), 
       col = c("steelblue", "firebrick"), pch = 19, lty = 1)
```

#### Difference-in-Difference

```{r}
C1 <- mean( dat$crimes[ dat$year == 1994 & dat$group == "control" ] )
C2 <- mean( dat$crimes[ dat$year == 1995 & dat$group == "control" ] )
C3 <- mean( dat$crimes[ dat$year == 1996 & dat$group == "control" ] )
C4 <- mean( dat$crimes[ dat$year == 1997 & dat$group == "control" ] )

T1 <- mean( dat$crimes[ dat$year == 1994 & dat$group == "treat" ] )
T2 <- mean( dat$crimes[ dat$year == 1995 & dat$group == "treat" ] )
T3 <- mean( dat$crimes[ dat$year == 1996 & dat$group == "treat" ] )
T4 <- mean( dat$crimes[ dat$year == 1997 & dat$group == "treat" ] )

dd.C1 <- mean( dat$crimes[ dat$post == 0 & dat$group == "control" ] )
dd.C2 <- mean( dat$crimes[ dat$post == 1 & dat$group == "control" ] )

dd.T1 <- mean( dat$crimes[ dat$post == 0 & dat$group == "treat" ] )
dd.T2 <- mean( dat$crimes[ dat$post == 1 & dat$group == "treat" ] )

plot( ts1$time, ts1$ave, type="b", pch=19, col="steelblue", bty="n",
      xlab="Time (months)", ylab="Number of Crimes", 
      main="Difference-in-Difference Group Means",
      xlim=c(0,50), ylim=c(60,225) )
points( ts2$time, ts2$ave, type="b", pch=19, col="firebrick" ) 
abline( v=c(12,36,48), lty=2, col="gray" )
abline( v=24.5, lty=3, col="gray", lwd=3 )
abline( h=c(y.24,y.25), lty=3, col="firebrick" )

points( c(6.5,18.5,30.5,42.5), c(C1,C2,C3,C4), 
        pch=19, col=adjustcolor("steelblue",alpha=0.5), cex=2 )
points( c(6.5,18.5,30.5,42.5), c(T1,T2,T3,T4), 
        pch=19, col=adjustcolor("firebrick",alpha=0.5), cex=2 )

points( c(12.5,36.5), c(dd.C1,dd.C2+3), col="steelblue", pch=19, cex=4 )
points( c(12.5,36.5), c(dd.T1,dd.T2-3), col="firebrick", pch=19, cex=4 )
legend("topleft", legend = c("Control", "Treatment"), 
       col = c("steelblue", "firebrick"), pch = 19, lty = 1)
```

```{r, results = 'asis'}
# Diff-in-Diff Model
m1 <- lm(crimes ~ treat + post.treat + treat.x.post, data = dat)

stargazer(m1, 
#          type = "html", 
          type = "text",
          digits = 2,
          omit.stat = c("f", "rsq", "adj.rsq", "ser"),
          title = "Difference-In-Difference",
          intercept.bottom = FALSE,
          dep.var.labels = "Crimes",
          covariate.labels = c("Intercept (a0)", 
                               "Treatment Group (a1)",
                               "Post-Treatment Period (a2)", 
                               "Treat × Post (a3)"))
```

#### Interrupted Time Series

```{r}
d2 <- filter( dat, group == "treat" )
 
m2 <- lm( crimes ~ time + post.treat + time.since, data=d2 )

plot( d2$time, d2$crimes, pch=19, 
      col=adjustcolor("firebrick",alph=0.5), cex=1.5, bty="n",
      xlab="Time (months)", 
      ylab="Count of Crimes Per Quadrant", 
      xlim=c(0,50) )
abline( v=c(12,36,48), lty=2, col="gray" )
abline( v=24.5, lty=3, col="gray", lwd=3 )
points( d2$time, m2$fitted.values, pch=19, type="l", lwd=3 )
```

```{r, results = 'asis' }
# Time Series Model
d2 <- filter(dat, group == "treat")
m2 <- lm(crimes ~ time + post.treat + time.since, data = d2)

stargazer(m2, 
#          type = "html",
          type = "text",
          digits = 2,
          omit.stat = c("f", "rsq", "adj.rsq", "ser"),
          title = "Interrupted Time Series",
          intercept.bottom = FALSE,
          dep.var.labels = "Crimes",
          covariate.labels = c("Intercept (b0)", 
                               "Time (b1)",
                               "Post-Treatment (b2)", 
                               "Time Since Treatment (b3)"))
```



#### Q4a: What is the size of the program impact according to the DID model? Report the coefficient name, value, and significance. (5 points)

**Type your answer here, within the asterisks**

#### Q4b: What is the size of the immediate program impact according to the ITS model? Report the coefficient name, value, and significance. (5 points)

**Type your answer here, within the asterisks**

#### Q4c: How are the post-treatment coefficients conceptually different in these two models (a2 vs b2)? (5 points)

**Type your answer here, within the asterisks**

#### Q4d: Which coefficient in each model captures the secular trend (changes we expect independent of treatment)? (5 points)

**Type your answer here, within the asterisks**

#### Q4e: Do the two models make similar predictions regarding secular trends? Show your math. (5 points)
  > **Note:** One model's secular trend is aggregated, the other is not. You will need to do some math to get the secular trends to represent the unit of time so you can accurately compare them.

```{r}
# Your code here
```

**Type your answer here, within the asterisks**

---

### BONUS Question (10 points)

#### BQ: Why does the DID model report a smaller program impact (a3) than the ITS immediate effect (b2)? (10 points)

**Type your answer here, within the asterisks**

---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header
* Name your files according to the convention: `Lab-##-LastName.Rmd`
* Show your solution and include your code
* Format your regression tables using `stargazer`
* **Bold your answers**
* Do **NOT** print excessive output (like a full data set)
* Follow appropriate style guidelines (spaces between arguments, etc.)

*Note: Platforms like Canvas sometimes disallow HTML file submissions when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*


```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```