---
title: "Lab 02: Difference-In-Difference"
subtitle: "PAF 512 - Advanced Regression"
author: "Lab Solutions"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>
```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **4 sets of questions** to be answered. You can earn up to **100 points + bonus questions**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each questions
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---

## The Policy Problem

This lab is loosely based on Branas et al. (2011), entitled "A Difference-in-Differences Analysis of Health, Safety, and Greening Vacant Urban Space".

The 'broken windows' theory suggests that vacant lots offer refuge to criminal and other illegal activity and visibly symbolize that a neighborhood has deteriorated. To prevent these problems, **City A** has decided to implement a public program that will "green" some of the city's vacant lots in 4 different neighborhoods. 'Greening' includes removing trash and debris and planting grass and trees to create a park.

Three years after implementation, the city wants to know whether the program has been successful. They propose comparing lots that have been 'greened' (treatment) to lots that are still vacant but could have been chosen for greening (control).

> **Research Question:** Does greening vacant lots decrease vandalism in City A?

> **Hypothesis:** Greening vacant lots will decrease vandalism in City A.

---

## Data

Data on vandalism acts were reported by the local police and include the total number of calls about vandalism acts received during the year before and after the greening from households near the vacant lots.

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/diff-in-diff-lab.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)
head(data)
```

### Variable Descriptions

| Variable Name | Description |
|---------------|-------------|
| Vandalism | Number of calls about vandalism acts near the vacant lots |
| Group | Treatment (=1) and control group (=0) |
| Post | Observation before treatment (=0) and after treatment (=1) |

### Load Packages

```{r}
# Install/Load Packages
# install.packages("stargazer") 
# install.packages("dplyr")
# install.packages("knitr")
library(wooldridge)   # Contains the injury dataset
library(stargazer)    # Clean regression tables
library(dplyr)        # Data manipulation
library(knitr)        # kable function for tables and inline code
```

---

## Lab Questions

### Question 1: The DID Model

#### Q1: Write down the difference-in-difference formula and build the model using the variables in our dataset.
  > **Note: **Write the formula in the asterisks, build the model using lm() in the code chunk. Extra kudos if you utilize LaTeX for your formula as shown in the codethrough.
  
```{r}
# Either create an interaction term, or simply interact the variables in your model. 
data$Group_Post <- data$Group * data$Post

# Run DID model (Group_Post = Group:Post = Group * Post)
DID_model <- lm(Vandalism ~ Group + Post + Group_Post, data = data)

# $$Vandalism = \beta_0 + \beta_1 \cdot Group + \beta_2 \cdot Post + \beta_3 \cdot (Group \times Post) + \epsilon$$
```

**$$Vandalism = \beta_0 + \beta_1 \cdot Group + \beta_2 \cdot Post + \beta_3 \cdot (Group \times Post) + \epsilon$$**

---

### Question 2: Model Estimation and Interpretation

#### Q2a: Run the DID regression and present results using stargazer.

```{r, results = "asis"}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Display results
stargazer(DID_model,
          #type = "text",
          type = "html",
          dep.var.labels = "Vandalism Calls",
          covariate.labels = c("Constant (β0)",
                               "Group (β1)", ## Intercept.bottom = FALSE
                               "Post (β2)",
                               "DID Estimator (β3)"),
          omit.stat = "all",
          digits = 2,
          intercept.bottom = FALSE) # intercept moved from bottom to top, requires strict ordering in the covariate.labels argument

```


#### Q2b: What is the average number of vandalism calls in the control group before the treatment? Which coefficient provides this information? (5 points)

**The average number of vandalism calls in the control group before treatment is approximately $2,241$ calls. This is provided by the Constant (**$\beta_0$**) coefficient, which represents the mean outcome for the reference group (Control, Pre-treatment).**

#### Q2c: Did the treatment and control groups differ significantly in vandalism calls before the intervention? Which coefficient tells us this, and what is its value and significance? (10 points)

**The Group coefficient (**$\beta_1$**) tells us whether the groups differed before treatment. The value is $26.26$, indicating that the treatment group had approximately $26$ more vandalism calls than the control group before the intervention. This coefficient is not statistically significant (p = 0.22).**

#### Q2d: Do treatment and control groups need to be identical in the pre-treatment period for valid DID inference? Explain why or why not. (10 points)

**No, treatment and control groups do NOT need to be identical in the pre-treatment period. The DID design accounts for baseline differences between groups through the** $\beta_1$ **coefficient. What matters is the parallel trends assumption—that both groups would have followed similar trends over time in the absence of treatment. The DID estimator captures the additional change in the treatment group beyond what we’d expect based on the control group’s trend, regardless of their starting levels.**

#### Q2e: Does the number of vandalism calls in the control group change after the intervention? Report the coefficient, whether it increases or decreases, and its significance. (5 points)

**The Post coefficient (**$\beta_2$**) = $318.08$ tells us the control group’s change over time. Vandalism calls in the control group increased by approximately 318 calls from the pre to post period. This coefficient is statistically significant (p < .01). This represents the secular trend—changes that occurred independent of the greening intervention.**

#### Q2f: What is the average number of vandalism calls in the control group after the intervention? Show the formula and calculation. (5 points)
  > Feel free to only type the answer, code is optional here.
  
```{r}
# Control group post-treatment mean = β0 + β2
b0 <- coef(DID_model)[1] 
b2 <- coef(DID_model)[3]

control_post <- b0 + b2

cat(
  "Control Group Post-Treatment Mean = β0 + β2\n",
  "= ", round(b0, 2), " + ", round(b2, 2), "\n",
  "= ", round(control_post, 2), " vandalism calls\n")
```


**The average number of vandalism calls in the control group after the intervention is $2,559$ calls, calculated as** $\beta_0$ **+ ** $\beta_2$**.**

#### Q2g: Does greening help reduce vandalism? Report the coefficient name, value, significance, and interpret its real-world meaning. (10 points)

**The DID Estimator (** $\beta_3$**) = $-953.1$ tells us the treatment effect. This coefficient is statistically significant (p < .01).**

**Real-world interpretation: Yes, greening appears to help reduce vandalism. After accounting for the secular trend, the greened lots experienced approximately $953$ fewer vandalism calls than we would have expected. This suggests the greening program successfully reduced vandalism.**

#### Q2h: What is the average number of vandalism calls in the treatment group after the intervention? Show the formula and calculation. (5 points)
  > Feel free to only type the answer, code is optional here.

```{r}
# Treatment group post-treatment mean = β0 + β1 + β2 + β3
b1 <- coef(DID_model)[2]
b3 <- coef(DID_model)[4]

treatment_post <- b0 + b1 + b2 + b3

cat("Treatment Group Post-Treatment Mean = β0 + β1 + β2 + β3\n",
"= ", round(b0, 2), " + ", round(b1, 2), " + ", round(b2, 2), " + ", round(b3, 2), "\n",
"= ", round(treatment_post, 2), " vandalism calls")
```

**The average number of vandalism calls in the treatment group after the intervention is $1,632$ calls, calculated as ** $\beta_0$ **+ **$\beta_1$ **+ ** $\beta_2$ **+ ** $\beta_3$**.**

---

### Question 3: Counterfactual Analysis

#### Q3a: Write the counterfactual formula using beta coefficients. What does the counterfactual represent? (5 points)

$$Counterfactual = \beta_0 + \beta_1 + \beta_2$$

**The counterfactual represents the expected number of vandalism calls in the treatment group (greened lots) if the intervention had NOT occurred. It is constructed by taking the treatment group’s baseline (**$\beta_0$ **+ **$\beta_1$**) and adding the secular trend observed in the control group (**$\beta_2$**).**

#### Q3b: Calculate the counterfactual value. How many vandalism calls would we expect if the intervention had not occurred? (5 points)
  > Feel free to only type the answer, code is optional here.

```{r}
# Calculate counterfactual
counterfactual <- b0 + b1 + b2

cat("Counterfactual = β0 + β1 + β2\n",
"= ", round(b0, 2), " + ", round(b1, 2), " + ", round(b2, 2), "\n",
"= ", round(counterfactual, 2), " vandalism calls\n")
```

**The counterfactual is $2,585$ vandalism calls. This means that if the greening program had not been implemented, we would have expected approximately $2,585$ vandalism calls in the treatment areas, based on their pre-treatment level and the trend observed in the control group.**

#### Q3c: What is the main assumption behind the counterfactual in a DID design? Why is this assumption important? (5 points)

**The main assumption is the parallel trends assumption. This assumption states that, in the absence of treatment, the treatment and control groups would have followed similar trends over time. This is critical because the entire DID logic relies on using the control group’s change to estimate what would have happened to the treatment group without the intervention. If the groups were already trending differently before the intervention, our counterfactual would be wrong and the treatment effect estimate would be biased.**

---

### Question 4: Comparing DID and ITS Approaches

Now let's compare two different approaches to analyzing policy interventions using a different dataset with monthly observations.

```{r}
## Load and preview the data
URL2 <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/time-series-diff-n-diff-comparison.csv"
dat <- read.csv(URL2, stringsAsFactors = FALSE)
rbind(head(dat, 3), tail(dat, 3))
```

This dataset contains 48 months of data. The vacant lot clean-up occurred between months 24 and 25.

#### Visualize the data

```{r}

# Calculate averages by group and time
ts_control <- dat %>% 
  filter(group == "control") %>%
  group_by(time) %>% 
  summarize(ave = mean(crimes), .groups = "drop")

ts_treat <- dat %>% 
  filter(group == "treat") %>%
  group_by(time) %>% 
  summarize(ave = mean(crimes), .groups = "drop")

# Plot
plot(ts_control$time, ts_control$ave, type = "b", pch = 19, col = "steelblue",
     xlab = "Time (months)", ylab = "Number of Crimes",
     main = "Crime Rates Over Time by Group",
     xlim = c(0, 50), ylim = c(60, 225), bty = "n")
points(ts_treat$time, ts_treat$ave, type = "b", pch = 19, col = "firebrick")
abline(v = 24.5, lty = 3, col = "gray", lwd = 3)

legend("topleft", legend = c("Control", "Treatment"), 
       col = c("steelblue", "firebrick"), pch = 19, lty = 1)
```
```{r}
ts1 <- 
  dat %>% 
  filter( group == "control" ) %>%
  group_by( time ) %>% 
  summarize( ave=mean( crimes ) )

ts2 <- 
  dat %>% 
  filter( group == "treat" ) %>%
  group_by( time ) %>% 
  summarize( ave=mean( crimes ) )

y.24 <- ts2$ave[24]
y.25 <- ts2$ave[25]
diff <- y.24 - y.25 

plot( ts1$time, ts1$ave, type="b", pch=19, col="steelblue", bty="n",
      xlab="Time (months)", ylab="Number of Crimes", 
      xlim=c(0,50), ylim=c(60,225) )
points( ts2$time, ts2$ave, type="b", pch=19, col="firebrick" ) 
abline( v=c(12,36,48), lty=2, col="gray" )
abline( v=24.5, lty=3, col="gray", lwd=3 )
abline( h=c(y.24,y.25), lty=3, col="firebrick" )

arrows( x0=38, y0=y.24, y1=y.25, col="firebrick", code=3 )
text( x=40, y=mean(c(y.24,y.25)), 
      paste0( "Diff ~ ", round(diff,0) ), 
      col="firebrick", pos=4, cex=1.4 )
legend("topleft", legend = c("Control", "Treatment"), 
       col = c("steelblue", "firebrick"), pch = 19, lty = 1)
```

#### Difference-in-Difference

```{r}
C1 <- mean( dat$crimes[ dat$year == 1994 & dat$group == "control" ] )
C2 <- mean( dat$crimes[ dat$year == 1995 & dat$group == "control" ] )
C3 <- mean( dat$crimes[ dat$year == 1996 & dat$group == "control" ] )
C4 <- mean( dat$crimes[ dat$year == 1997 & dat$group == "control" ] )

T1 <- mean( dat$crimes[ dat$year == 1994 & dat$group == "treat" ] )
T2 <- mean( dat$crimes[ dat$year == 1995 & dat$group == "treat" ] )
T3 <- mean( dat$crimes[ dat$year == 1996 & dat$group == "treat" ] )
T4 <- mean( dat$crimes[ dat$year == 1997 & dat$group == "treat" ] )

dd.C1 <- mean( dat$crimes[ dat$post == 0 & dat$group == "control" ] )
dd.C2 <- mean( dat$crimes[ dat$post == 1 & dat$group == "control" ] )

dd.T1 <- mean( dat$crimes[ dat$post == 0 & dat$group == "treat" ] )
dd.T2 <- mean( dat$crimes[ dat$post == 1 & dat$group == "treat" ] )

plot( ts1$time, ts1$ave, type="b", pch=19, col="steelblue", bty="n",
      xlab="Time (months)", ylab="Number of Crimes", 
      main="Difference-in-Difference Group Means",
      xlim=c(0,50), ylim=c(60,225) )
points( ts2$time, ts2$ave, type="b", pch=19, col="firebrick" ) 
abline( v=c(12,36,48), lty=2, col="gray" )
abline( v=24.5, lty=3, col="gray", lwd=3 )
abline( h=c(y.24,y.25), lty=3, col="firebrick" )

points( c(6.5,18.5,30.5,42.5), c(C1,C2,C3,C4), 
        pch=19, col=adjustcolor("steelblue",alpha=0.5), cex=2 )
points( c(6.5,18.5,30.5,42.5), c(T1,T2,T3,T4), 
        pch=19, col=adjustcolor("firebrick",alpha=0.5), cex=2 )

points( c(12.5,36.5), c(dd.C1,dd.C2+3), col="steelblue", pch=19, cex=4 )
points( c(12.5,36.5), c(dd.T1,dd.T2-3), col="firebrick", pch=19, cex=4 )
legend("topleft", legend = c("Control", "Treatment"), 
       col = c("steelblue", "firebrick"), pch = 19, lty = 1)
```

```{r, results = "asis"}
# Diff-in-Diff Model
m1 <- lm(crimes ~ treat + post.treat + treat.x.post, data = dat)

stargazer(m1, 
          type = "html", 
          #type = "text",
          digits = 2,
          omit.stat = c("f", "rsq", "adj.rsq", "ser"),
          title = "Difference-In-Difference",
          intercept.bottom = FALSE,
          dep.var.labels = "Crimes",
          covariate.labels = c("Intercept (a0)", 
                               "Treatment Group (a1)",
                               "Post-Treatment Period (a2)", 
                               "Treat × Post (a3)"))
```

#### Interrupted Time Series

```{r}
d2 <- filter( dat, group == "treat" )
 
m2 <- lm( crimes ~ time + post.treat + time.since, data=d2 )

plot( d2$time, d2$crimes, pch=19, 
      col=adjustcolor("firebrick",alph=0.5), cex=1.5, bty="n",
      xlab="Time (months)", 
      ylab="Count of Crimes Per Quadrant", 
      xlim=c(0,50) )
abline( v=c(12,36,48), lty=2, col="gray" )
abline( v=24.5, lty=3, col="gray", lwd=3 )
points( d2$time, m2$fitted.values, pch=19, type="l", lwd=3 )
```

```{r, results = "asis"}
# Time Series Model
d2 <- filter(dat, group == "treat")
m2 <- lm(crimes ~ time + post.treat + time.since, data = d2)

stargazer(m2, 
          type = "html",
          #type = "text",
          digits = 2,
          omit.stat = c("f", "rsq", "adj.rsq", "ser"),
          title = "Interrupted Time Series",
          intercept.bottom = FALSE,
          dep.var.labels = "Crimes",
          covariate.labels = c("Intercept (b0)", 
                               "Time (b1)",
                               "Post-Treatment (b2)", 
                               "Time Since Treatment (b3)"))
```



#### Q4a: What is the size of the program impact according to the DID model? Report the coefficient name, value, and significance. (5 points)

**The program impact in the DID model is captured by the Treat × Post coefficient (**$\alpha_3$**) = $-23.09$. This indicates that the greening program reduced crimes by approximately $23$ crimes compared to what we would expect based on the control group’s trend. This coefficient is statistically significant (p < 0.01).**

#### Q4b: What is the size of the immediate program impact according to the ITS model? Report the coefficient name, value, and significance. (5 points)

**The immediate program impact in the ITS model is captured by the Post-Treatment coefficient (**$\beta_2$**) = $-36.43$. This indicates an immediate decrease of approximately $36$ crimes when the treatment occurred. This coefficient is statistically significant (p < 0.01).**

#### Q4c: How are the post-treatment coefficients conceptually different in these two models (a2 vs b2)? (5 points)

**These coefficients measure fundamentally different things:**

  - $\alpha_2$ **(Post in DID):** Represents the secular trend—how the CONTROL group changed from pre to post period. This captures changes that happened to everyone regardless of treatment.

  - $\beta_2$ **(Post in ITS):** Represents the immediate treatment effect—the level shift in the TREATMENT group at the moment of intervention. This is the “jump” up or down when treatment begins.

**In DID, **$\alpha_2$ **is about the control group’s change; in ITS, **$\beta_2$ **is about the treatment group’s immediate change.**

#### Q4d: Which coefficient in each model captures the secular trend (changes we expect independent of treatment)? (5 points)

**DID Model: The secular trend is captured by **$\alpha_2$ **(Post-Treatment Period) = $64.85$. This coefficient shows how the control group changed over time, which we assume reflects trends affecting everyone.**

**ITS Model: The secular trend is captured by** $\beta_1$ **(Time) = $2.67$. This coefficient represents the pre-treatment slope—the rate of change before any intervention occurred.**

#### Q4e: Do the two models make similar predictions regarding secular trends? Show your math. (5 points)
  > **Note:** One model's secular trend is aggregated, the other is not. You will need to do some math to get the secular trends to represent the unit of time so you can accurately compare them.

```{r}
# DID secular trend: a2 represents total change over the period
did_secular <- coef(m1)[3]

# ITS secular trend: b1 represents change per month
# Total change over 24 months (pre-treatment period)
its_secular_per_month <- coef(m2)[2]
its_secular_total <- its_secular_per_month * 24

cat(
  "DID Secular Trend (a2):", round(did_secular, 2), "crimes (total change)\n",
  "ITS Secular Trend (b1):", round(its_secular_per_month, 2), "crimes per month\n",
  "ITS over 24 months:", round(its_secular_per_month, 2), "× 24 =", round(its_secular_total, 2), "crimes\n")

```

**The DID model estimates a secular trend of $64.85$ crimes (total change in control group from pre to post).**

**The ITS model estimates a secular trend of $2.67$ crimes per month. Over 24 months, this equals $63.99$ crimes.**

**The estimates are reasonably similar, which makes sense because both models are trying to capture the same underlying trend using different approaches.**

---

### BONUS Question (10 points)

#### BQ: Why does the DID model report a smaller program impact (a3) than the ITS immediate effect (b2)? (10 points)

**The DID estimate (**$\alpha_3$**) is smaller because it accounts for what happened to the control group over the same period. Looking at the graph, we can see that BOTH groups increased over time (secular trend). The DID isolates only the ADDITIONAL change in the treatment group beyond this common trend.**

**The ITS estimate (**$\beta_2$**) captures the total immediate change in the treatment group, which includes both the treatment effect AND any secular changes that happened to be occurring at that time. Without a control group, ITS cannot separate these.**

**This illustrates why having a control group (DID) can provide cleaner causal estimates when one is available.**


---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header
* Name your files according to the convention: `Lab-##-LastName.Rmd`
* Show your solution and include your code
* Format your regression tables using `stargazer`
* **Bold your answers**
* Do **NOT** print excessive output (like a full data set)
* Follow appropriate style guidelines (spaces between arguments, etc.)

*Note: Platforms like Canvas sometimes disallow HTML file submissions when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```