---
title: "Lab 03: Fixed Effects Models"
subtitle: "PAF 512 - Advanced Regression"
author: "Lab Solutions"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>
```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **4 sets of questions** to be answered. You can earn up to **100 points + 10 bonus points**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each question
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---

## The Policy Problem

> **Research Question:** Do beer taxes reduce car accidents?

Several studies have examined how state alcohol control policies affect traffic safety. Beer taxes are one mechanism states use to control alcohol consumption—higher taxes increase prices, which decrease consumption. Lower alcohol consumption is expected to reduce drunk driving and therefore accidents, especially at night.

> **Hypothesis:** Higher beer taxes will be associated with fewer car accidents.

---

## Data

We are using panel data from 7 southern US states observed across 7 years (2010-2016). This structure—repeated observations of the same units over time—is essential for fixed effects analysis.

### Load Packages

```{r}
# Install packages if needed
# install.packages("stargazer")
# install.packages("plm")
# install.packages("dplyr")

library(stargazer)  # For nice regression tables
library(plm)        # For panel data models
library(dplyr)      # For data wrangling
```

### Load Data

```{r}
# Load the panel dataset
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/beer-tax-fixed-effects.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)

# Preview the data structure
head(data[, c("state", "year", "taxes", "accidents")], 14)
```

### Variable Descriptions

| Variable | Description |
|----------|-------------|
| `state` | State identifier (1-7) |
| `year` | Year of observation (2010-2016) |
| `taxes` | Beer tax as proportion of cost (0 to 1) |
| `accidents` | Number of car accidents |

---

<div class="caution-note">

## A Note on Interpreting Taxes

Our `taxes` variable is scaled as a proportion from 0 to 1 (e.g., 0.05 = 5% tax).

In regression, a coefficient always represents the effect of a 1-unit increase in X. Since our scale is 0–1, a "1-unit increase" implies increasing the tax by 100 percentage points (going from 0% to 100%). This is mathematically correct but practically unrealistic.

To make our interpretations meaningful, we will interpret the effect at the 1-percentage-point level:

  * **The Math:** Since one-percentage point is $0.01$ on our scale, we simply multiply the coefficient by $0.01$
  
  * **Example:** If the coefficient is $-20$, that represents the effect of a full 0-to-1 jump. To find the effect of a one-percentage point increase, we calculate $-20 * 0.01 = -0.2$.
  
  * **Interpretation:** For every one-percentage point increase in the beer tax, we expect $0.2$ fewer accidents per year.
  
</div>

## Lab Questions

### Question 1: Pooled OLS Model

Your colleague begins by analyzing the data using a pooled OLS model, which ignores the panel structure and treats all observations as independent.

#### Q1a: Write the formula for a pooled OLS regression model predicting accidents from taxes. (5 points)

$$Accidents = \beta_0 + \beta_1 \cdot Taxes + \epsilon$$

#### Q1b: Run the pooled OLS model and present your results using stargazer. (10 points)

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), 
# but remember to set type = "html" when you are ready to knit and submit your lab.

# Run the pooled OLS model
pooled_ols <- lm(accidents ~ taxes, data = data)

# Display results
stargazer(pooled_ols,
          type = "html",
          #type = "text"
          title = "Pooled OLS Model",
          dep.var.labels = "Car Accidents",
          covariate.labels = c("Pooled Average", "Beer Taxes"),
          omit.stat = "all",
          digits = 2,
          intercept.bottom = FALSE)
```

#### Q1c: Interpret the coefficient on `taxes`. In your answer, report the coefficient value, its statistical significance, and explain what the number means in real-world terms. Based on this model, do beer taxes appear to reduce car accidents? (10 points)

**The coefficient on taxes is $760.12$ and is NOT statistically significant (t = 1.54, p > 0.05). This suggests that one-percentage point increase in the beer tax (e.g., going from 0% to 1% of cost) is associated with approximately $7.6$ MORE car accidents each year. This is the opposite of what we hypothesized—according to this model, higher taxes appear to INCREASE accidents. However, this coefficient is not statistically significant, and this is a pooled model that ignores state-level differences, so we should be cautious about drawing any conclusions.**

---

### Question 2: Fixed Effects Models

You suspect the pooled OLS results may be biased due to Simpson's Paradox—states with higher baseline accident rates might also have higher beer taxes (perhaps because they implemented taxes in response to accident problems). You propose using fixed effects to control for these state-level differences.

#### Q2a: Write the formula for a fixed effects model using dummy variables for each state. (5 points)

  > **Note:** Use $S_1, S_2, ... S_7$ to represent state dummy variables.

$$Accidents = \beta_1 \cdot Taxes + \alpha_1 S_1 + \alpha_2 S_2 + \alpha_3 S_3 + \alpha_4 S_4 + \alpha_5 S_5 + \alpha_6 S_6 + \alpha_7 S_7 + \epsilon$$

$$ Y_{it} = \beta_1 X_{it} + \alpha_i + e_{it} $$


#### Q2b: Looking at your formulas from Q1a and Q2a, what is the main difference between a pooled OLS model and a fixed effects model? (5 points)

**The main difference is that the fixed effects model includes a separate intercept (dummy variable) for each state, while the pooled OLS has only one intercept for all observations. The fixed effects model controls for all time-invariant characteristics of each state by giving each state its own baseline level of accidents.**

#### Q2c: Run a fixed effects model using `lm()` with state dummy variables. Use `-1` in your formula to suppress the global intercept so that each state gets its own intercept coefficient. Present your results using stargazer. (5 points)
> **Note:** The `-1` removes the default intercept, allowing us to see the intercept for every state rather than having one reference category.

```{r, results='asis'}
# Run fixed effects using lm() with state dummies
# The -1 removes the global intercept so we get an intercept for each state
fe_lm <- lm(accidents ~ taxes + factor(state) - 1, data = data)

# Display results
stargazer(fe_lm,
          type = "html",
          #type = "text"
          title = "Fixed Effects Model (lm with dummies)",
          dep.var.labels = "Car Accidents",
          covariate.labels = c("Beer Taxes", 
                               "State 1", "State 2", "State 3", 
                               "State 4", "State 5", "State 6", "State 7"),
          omit.stat = "all",
          digits = 2)
```

#### Q2d: Now run the same fixed effects model using the `plm()` function with `model = "within"`. Present both models (Q2c and Q2d) side-by-side in a single stargazer table. (5 points)
> **Note:** The `plm()` function is the professional standard for panel data analysis. It produces the same coefficient on taxes but handles the state effects differently in its output.

```{r, results='asis'}
# Run fixed effects using plm()
# index tells plm which variable identifies the groups (states)
fe_plm <- plm(accidents ~ taxes, 
              data = data, 
              index = c("state"), 
              model = "within")

# Display both models side by side
stargazer(fe_lm, fe_plm,
          type = "html",
          #type = "text"
          title = "Fixed Effects: lm() vs plm()",
          dep.var.labels = "Car Accidents",
          column.labels = c("FE with lm()", "FE with plm()"),
          covariate.labels = c("Beer Taxes", 
                               "State 1", "State 2", "State 3", 
                               "State 4", "State 5", "State 6", "State 7"),
          omit.stat = "all",
          digits = 2)
```

#### Q2e: Compare the coefficient on `taxes` between the pooled OLS model (Q1b) and the fixed effects model. Report both coefficient values, their significance levels, and explain how controlling for state-level differences changed our conclusions about the effect of beer taxes on accidents. (10 points)

**In the pooled OLS model, the taxes coefficient was $760.12$ (not statistically significant, t = 1.54), suggesting higher taxes might increase accidents but we can't be confident in this result. In the fixed effects model, the taxes coefficient is $-31.37$ (statistically significant, t = -2.28, p < 0.05), suggesting higher taxes DECREASE accidents by approximately $0.31$ accidents annually per 1-percentage point increase in the tax rate.**

**Controlling for state-level differences reversed our conclusion AND made it statistically significant. This is Simpson's Paradox in action—states with more accidents at baseline (like States 1, 2, 3, and 5 with ~2,000+ accidents) have different tax levels than states with fewer accidents (like State 4 with ~900 accidents). Once we control for each state's baseline accident level, we see that WITHIN states, higher taxes are associated with fewer accidents. This supports our original hypothesis.**

---

### Question 3: Model Diagnostics & Interpretation

#### Q3a: Use the `pFtest()` function to formally test whether the fixed effects model is preferred over the pooled OLS model. Report the test statistic, p-value, and state your conclusion about which model is more appropriate for this data. (10 points)
> **Note:** The null hypothesis of `pFtest()` is that the pooled OLS model is adequate (i.e., fixed effects are not needed).

```{r}
# Run the F-test comparing FE to pooled OLS
pFtest(fe_plm, pooled_ols)
```

**The F-test produces an F-statistic of $11080.37$ with a p-value essentially equal to zero (p < 2.2e-16). We reject the null hypothesis that pooled OLS is adequate. This means the fixed effects model is significantly better than pooled OLS—there are meaningful state-level differences that must be accounted for. We should use the fixed effects model.**

#### Q3b: Use the `fixef()` function to retrieve the state intercepts from your `plm()` model. What does State 1's intercept value represent? (5 points)

```{r}
# Retrieve the fixed effects (state intercepts) from plm model
fixef(fe_plm)
```

**State 1's intercept is $2090.57$. This represents the predicted number of car accidents in State 1 when the beer tax is equal to zero. It captures State 1's baseline level of accidents after accounting for all time-invariant characteristics of that state (geography, culture, road infrastructure, etc.).**

#### Q3c: Using your `lm()` model from Q2c, what is the intercept value for State 4? Explain what this value tells us about car accidents in State 4. (10 points)

**State 4's intercept is $930.75$. This represents the predicted number of accidents in State 4 when the beer tax equals zero. Comparing to State 1's intercept of $2090.57$, State 4 has a dramatically lower baseline level of accidents—approximately $1,160$ fewer accidents than State 1, all else being equal. This large difference reflects time-invariant characteristics of State 4 (perhaps smaller population, different road infrastructure, or geographic factors) that make it much less prone to accidents than other states in the dataset.**


---

### Question 4: Understanding Fixed Effects Assumptions

#### Q4a: Hypothetically, suppose State 7 kept its beer tax rate exactly the same (constant) for all 7 years of the study. How would this state contribute to the estimation of the taxes coefficient in the Fixed Effects model? Would it help us determine if taxes reduce accidents? (5 points)

**If State 7 never changed its tax rate, it would contribute zero information to the estimate of the tax coefficient. Fixed Effects models rely entirely on within-state variation (changes over time). If $X$ does not vary within a state, that state effectively 'drops out' of the slope estimation for that variable. We would be estimating the effect based only on the other 6 states that did change their taxes.**

#### Q4b: Your colleague wants to add the following variables to the model. For each variable, state whether it SHOULD or SHOULD NOT be included in a fixed effects model and explain why. (10 points)

1. **State geographical location (north, south, east, west)**
2. **Annual state unemployment rate**
3. **State form of government (commonwealth, Bicameral, etc.)**

**1. State geographical region: SHOULD NOT be included. Geographic region is a time-invariant characteristic—a state's region doesn't change over time. Fixed effects already control for this through the state intercepts. Including it would create perfect multicollinearity with the state dummies.**

**2. Annual state unemployment rate: SHOULD be included. Unemployment rates change over time within states, so this is a time-varying variable. Fixed effects can estimate its relationship with accidents.**

**3. State form of government: SHOULD NOT be included. The form of government is essentially time-invariant (or changes very rarely). Like geographic region, this is absorbed by the state fixed effects.**

#### Q4c: Based on your answer above, what is the key limitation of fixed effects models for researchers? What types of research questions CANNOT be answered using fixed effects? (5 points)

**The key limitation is that fixed effects cannot estimate the effect of time-invariant variables. Any characteristic that doesn't change over time within a group is absorbed by (perfectly collinear with) the group intercepts. This means researchers cannot use fixed effects to answer questions like "Do southern states have more accidents than northern states?" or "Does a particular form of government lead to better outcomes?" To answer such questions, you would need a different approach (like random effects or cross-sectional analysis), though those come with their own assumptions and limitations.**

---

### Bonus: De-Meaning Approach (10 points)

Fixed effects can also be estimated by "de-meaning" the data—subtracting each group's mean from every observation. This removes all between-group variation, leaving only within-group variation.

#### BQ1: De-mean both the `taxes` and `accidents` variables by state, then run an OLS regression on the de-meaned data. Present your results using stargazer. (5 points)

```{r, results='asis'}
# De-mean the data by state
# For each observation, subtract that state's mean
data_demeaned <- data %>%
  group_by(state) %>%
  mutate(
    taxes_demeaned = taxes - mean(taxes),
    accidents_demeaned = accidents - mean(accidents)
  ) %>%
  ungroup()

# Run OLS on de-meaned data
demeaned_ols <- lm(accidents_demeaned ~ taxes_demeaned, data = data_demeaned)

# Display results
stargazer(demeaned_ols,
          type = "html",
          #type = "text"
          title = "De-meaned OLS Model",
          dep.var.labels = "Accidents (de-meaned)",
          covariate.labels = c("Beer Taxes (de-meaned)"),
          keep = "taxes_demeaned",
          omit.stat = "all",
          digits = 2,
          intercept.bottom = FALSE)
```

#### BQ2: Compare the coefficient on taxes from the de-meaned OLS model to the coefficient from your fixed effects models. Are they the same? Explain why this result makes sense. (5 points)

**Yes, the coefficient on taxes from the de-meaned OLS ($-31.37$) is identical to the coefficient from both fixed effects models. This makes sense because de-meaning and adding dummy variables are mathematically equivalent ways of removing between-group variation. When we subtract each state's mean, we eliminate all the stable differences between states—exactly what the state dummy variables do. Both approaches isolate the within-state relationship between taxes and accidents, which is why they produce the same coefficient.**

---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-##-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Canvas sometimes disallows HTML file submissions when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Commons License:
(CC BY-NC-SA 4.0)

PAF 512 - Advanced Regression
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```
