---
title: "Lab 03: Fixed Effects Models"
subtitle: "PAF 512 - Advanced Regression"
author: "Insert Your First and Last Name Here"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>
```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **4 sets of questions** to be answered. You can earn up to **100 points + 10 bonus points**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- **Bold your answers**
- Submit your R Markdown file and knitted HTML

---

## The Policy Problem

> **Research Question:** Do beer taxes reduce car accidents?

Several studies have examined how state alcohol control policies affect traffic safety. Beer taxes are one mechanism states use to control alcohol consumption—higher taxes increase prices, which decrease consumption. Lower alcohol consumption is expected to reduce drunk driving and therefore accidents, especially at night.

> **Hypothesis:** Higher beer taxes will be associated with fewer car accidents.

---

## Data

We are using panel data from 7 southern US states observed across 7 years (2010-2016). This structure—repeated observations of the same units over time—is essential for fixed effects analysis.

### Load Packages

```{r}
# Install packages if needed
# install.packages("stargazer")
# install.packages("plm")
# install.packages("dplyr")

library(stargazer)  # For nice regression tables
library(plm)        # For panel data models
library(dplyr)      # For data wrangling
```

### Load Data

```{r}
# Load the panel dataset
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/beer-tax-fixed-effects.csv"
data <- read.csv(URL, stringsAsFactors = FALSE)

# Preview the data structure
head(data[, c("state", "year", "taxes", "accidents")], 14)
```

### Variable Descriptions

| Variable | Description |
|----------|-------------|
| `state` | State identifier (1-7) |
| `year` | Year of observation (2010-2016) |
| `taxes` | Beer tax as proportion of cost (0 to 1) |
| `accidents` | Number of car accidents |

---

<div class="caution-note">

## A Note on Interpreting Taxes

Our `taxes` variable is scaled as a proportion from 0 to 1 (e.g., 0.05 = 5% tax).

In regression, a coefficient always represents the effect of a 1-unit increase in X. Since our scale is 0–1, a "1-unit increase" implies increasing the tax by 100 percentage points (going from 0% to 100%). This is mathematically correct but practically unrealistic.

To make our interpretations meaningful, we will interpret the effect at the 1-percentage-point level:
  * **The Math:** Since one-percentage point is $0.01$ on our scale, we simply multiply the coefficient by $0.01$
  * **Example:** If the coefficient is $-20$, that represents the effect of a full 0-to-1 jump. To find the effect of a one-percentage point increase, we calculate $-20 * 0.01 = -0.2$.
  * **Interpretation:** For every one-percentage point increase in the beer tax, we expect $0.2$ fewer accidents per year.
  
</div>

## Lab Questions

### Question 1: Pooled OLS Model

Your colleague begins by analyzing the data using a pooled OLS model, which ignores the panel structure and treats all observations as independent.

#### Q1a: Write the formula for a pooled OLS regression model predicting accidents from taxes. (5 points)

**Type your answer here, within the asterisks. Extra kudos for using LaTeX.**

#### Q1b: Run the pooled OLS model and present your results using stargazer. (10 points)

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), 
# but remember to set type = "html" when you are ready to knit and submit your lab.

# Your code here
```

#### Q1c: Interpret the coefficient on `taxes`. In your answer, report the coefficient value, its statistical significance, and explain what the number means in real-world terms. Based on this model, do beer taxes appear to reduce car accidents? (10 points)

**Type your answer here, within the asterisks**

---

### Question 2: Fixed Effects Models

You suspect the pooled OLS results may be biased due to Simpson's Paradox—states with higher baseline accident rates might also have higher beer taxes (perhaps because they implemented taxes in response to accident problems). You propose using fixed effects to control for these state-level differences.

#### Q2a: Write the formula for a fixed effects model using dummy variables for each state. (5 points)
> **Note:** Use $S_1, S_2, ... S_7$ to represent state dummy variables.

**Type your answer here, within the asterisks. Extra kudos for using LaTeX.**

#### Q2b: Looking at your formulas from Q1a and Q2a, what is the main difference between a pooled OLS model and a fixed effects model? (5 points)

**Type your answer here, within the asterisks.**

#### Q2c: Run a fixed effects model using `lm()` with state dummy variables. Use `-1` in your formula to suppress the global intercept so that each state gets its own intercept coefficient. Present your results using stargazer. (5 points)
> **Note:** The `-1` removes the default intercept, allowing us to see the intercept for every state rather than having one reference category.

```{r, results='asis'}
# Run fixed effects using lm() with state dummies
# The -1 removes the global intercept so we get an intercept for each state

# Reminder: While completing your lab, set type = "text" for your stargazer(), 
# but remember to set type = "html" when you are ready to knit and submit your lab.

# Your code here
```

#### Q2d: Now run the same fixed effects model using the `plm()` function with `model = "within"`. Present both models (Q2c and Q2d) side-by-side in a single stargazer table. (5 points)
> **Note:** The `plm()` function is the professional standard for panel data analysis. It produces the same coefficient on taxes but handles the state effects differently in its output.

```{r, results='asis'}
# Run fixed effects using plm()
# index tells plm which variable identifies the groups (states)

# Reminder: While completing your lab, set type = "text" for your stargazer(), 
# but remember to set type = "html" when you are ready to knit and submit your lab.

# Your code here
```

#### Q2e: Compare the coefficient on `taxes` between the pooled OLS model (Q1b) and the fixed effects model. Report both coefficient values, their significance levels, and explain how controlling for state-level differences changed our conclusions about the effect of beer taxes on accidents. (10 points)

**Type your answer here, within the asterisks.**

---

### Question 3: Model Diagnostics & Interpretation

#### Q3a: Use the `pFtest()` function to formally test whether the fixed effects model is preferred over the pooled OLS model. Report the test statistic, p-value, and state your conclusion about which model is more appropriate for this data. (10 points)
> **Note:** The null hypothesis of `pFtest()` is that the pooled OLS model is adequate (i.e., fixed effects are not needed).

```{r}
# Your code here
```

**Type your answer here, within the asterisks.**

#### Q3b: Use the `fixef()` function to retrieve the state intercepts from your `plm()` model. What does State 1's intercept value represent? (5 points)

```{r}
# Your code here
```

**Type your answer here, within the asterisks.**

#### Q3c: Using your `lm()` model from Q2c, what is the intercept value for State 4? Explain what this value tells us about car accidents in State 4. (10 points)

**Type your answer here, within the asterisks.**


---

### Question 4: Understanding Fixed Effects Assumptions

#### Q4a: Hypothetically, suppose State 7 kept its beer tax rate exactly the same (constant) for all 7 years of the study. How would this state contribute to the estimation of the taxes coefficient in the Fixed Effects model? Would it help us determine if taxes reduce accidents? (5 points)

**Type your answer here, within the asterisks.**

#### Q4b: Your colleague wants to add the following variables to the model. For each variable, state whether it SHOULD or SHOULD NOT be included in a fixed effects model and explain why. (10 points)

1. **State geographical location (north, south, east, west)**
2. **Annual state unemployment rate**
3. **State form of government (commonwealth, Bicameral, etc.)**

**Type your answer here, within the asterisks.**

#### Q4c: Based on your answer above, what is the key limitation of fixed effects models for researchers? What types of research questions CANNOT be answered using fixed effects? (5 points)

**Type your answer here, within the asterisks.**

---

### Bonus: De-Meaning Approach

Fixed effects can also be estimated by "de-meaning" the data—subtracting each group's mean from every observation. This removes all between-group variation, leaving only within-group variation.

#### BQ1: De-mean both the `taxes` and `accidents` variables by state, then run an OLS regression on the de-meaned data. Present your results using stargazer. (5 points)

```{r, results='asis'}
# Reminder: While completing your lab, set type = "text" for your stargazer(), 
# but remember to set type = "html" when you are ready to knit and submit your lab.

# Your code here
```

#### BQ2: Compare the coefficient on taxes from the de-meaned OLS model to the coefficient from your fixed effects models. Are they the same? Explain why this result makes sense. (5 points)

**Type your answer here, within the asterisks.**

---

## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-##-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Platforms like BlackBoard and Canvas sometimes disallow you from submitting HTML files when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```