---
title: "Lab 01: Interrupted Time Series"
subtitle: "PAF 512 - Advanced Regression"
author: "Insert Your First and Last Name Here"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---
```{r, include = FALSE}
# The code below serves for improving the aesthetics of the lab document.
# Please do not modify the below code.
```

<style>
  /* --- ASU Theme Elements --- */
   
  /* Maroon Headers */
  h1, h2, h3, h4 {
    color: #8C1D40 !important;
  }
   
  /* Gold Links */
  a {
    color: #D4AF37 !important;
  }
   
  /* --- Cautionary Note Box --- */
  .caution-note {
    background-color: #fff3e0 !important; /* Light Orange/Cream background */
    border-left: 6px solid #d35400;        /* Burnt Orange border */
    color: #2c3e50;
    padding: 15px;
    margin-bottom: 20px;
    border-radius: 4px;
    box-shadow: 0 2px 4px rgba(0,0,0,0.1);
  }
  
  .caution-note strong {
    color: #d35400;
    font-size: 1.1em;
  }

  /* --- Table of Contents Styling --- */

  /* 1. Active State */
  .list-group-item.active, 
  .list-group-item.active:focus, 
  .list-group-item.active:hover {
    background-color: #8C1D40 !important;
    border-color: #8C1D40 !important;
    color: white !important;
  }

  /* 2. Subheaders (Level 2) */
  .tocify-subheader .list-group-item {
    background-color: #f9f9f9; 
    border-left: 4px solid #FFC627 !important; /* ASU Gold */
    font-size: 0.9em; 
    color: #555; 
    padding-left: 15px !important; 
  }

  /* 3. Sub-Subheaders (Level 3) */
  .tocify-subheader .tocify-subheader .list-group-item {
    border-left: 4px solid #8C1D40 !important; /* ASU Maroon */
    padding-left: 30px !important; 
  }

  /* 4. Hover Effects */
  .tocify-subheader .list-group-item:hover {
    background-color: #fffdf2 !important;
    color: #8C1D40 !important;
  }
  
  /* --- Code Block Formatting --- */
  pre {
    background-color: #f4f4f4 !important; 
    border: 1px solid #ddd !important;     
    color: #212529 !important;             
    font-weight: 500;                      
  }

  /* Comments */
  code span.co {
    color: #486a48 !important; 
    font-style: italic;        
    font-weight: normal;       
  }

  /* Strings */
  code span.st {
    color: #8C1D40 !important; 
    font-weight: bold;
  }
   
  /* Numbers and Keywords */
  code span.dv, code span.kw {
     color: #004d40 !important; 
  }
</style>
```{r, include= FALSE}
# The code above serves for improving the aesthetics of the lab document.
# Please do not modify the above code.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<div class="caution-note">
## AI & Coding Policy

* **Responsible AI Use:** Use AI to *support* your learning (debugging, explaining), not to *replace* it. Submitting code or answers generated by AI is not permitted.

* **Course Methods Required:** Please solve these problems using the specific R techniques taught in this module. AI may suggest shortcuts or different packages; avoid using these, as work relying on methodologies not found in the course materials will result in a loss of credit.
  
</div>

---

## Instructions

There are **4 sets of questions** to be answered. You can earn up to **100 points + bonus questions**. Points are indicated next to each question.

Remember to:

- Format your tables using `stargazer`
- Answer all parts of each questions
- Submit the R markdown file & HTML with all your code
- **Bold your answers**

---


## The Policy Problem

> **Research Question:** Does an increase in bus frequency affect bus ridership in City A?

Several cities aim to support public transportation and increase ridership. There are several ways to encourage residents to utilize public transportation more often; one of those is to modify the transit schedule so as to increase the frequency of buses in peak times and reduce waiting time.

City A has opted for this solution and starting from May 1st has implemented a new bus schedule which increases bus frequency. At the end of the year, the mayor wants to estimate whether the decision was effective and ask you to analyze one year of ridership data to test the following hypothesis:

> **Hypothesis:** An increase in bus frequency has a positive effect on bus ridership in City A.

--- 

## Data
Data on ridership were provided by the local transit agency. The variable Passengers represents the number of daily passengers on all buses in the city (in thousands).

  - **Collection Period:** January 1st to December 31st (365 days)
  
  - **Intervention Date:** May 1st (Day 121)


### Load Packages
```{r}
# Install packages, if needed.
  # install.packages("stargazer")
  # install.packages("lmtest")
  # install.packages("sandwich")
  # install.packages("dplyr")

# Load necessary libraries
  library(stargazer) # for nice regression tables
  library(lmtest)   # For Durbin-Watson test
  library(sandwich) # For Newey-West SEs
  library(dplyr) # For data manipulation
```

### Starting Data
The intervention was implemented on May 1st (Day 121). The variable Passengers represents the number of daily passengers (in thousands).

```{r}
# Load the raw data vector
passengers <- c(1328, 1407, 1425, 1252, 1287, 1353, 1301, 1294, 1336, 1371, 1408, 1326, 1364, 1295, 1320, 1260, 1347, 1316, 1287, 1292, 1259, 1349, 1274, 1365, 1317, 1341, 1316, 1313, 1285, 1369, 1309, 1446, 1422, 1397, 1358, 1310, 1294, 1373, 1161, 1320, 1376, 1335, 1382, 1455, 1374, 1267, 1318, 1370, 1297, 1391, 1269, 1341, 1238, 1391, 1296, 1260, 1330, 1447, 1296, 1389, 1278, 1319, 1333, 1372, 1325, 1299, 1299, 1312, 1352, 1355, 1404, 1317, 1330, 1325, 1368, 1311, 1310, 1242, 1247, 1366, 1401, 1282, 1298, 1301, 1341, 1353, 1398, 1352, 1300, 1442, 1365, 1411, 1360, 1100, 1334, 1336, 1274, 1303, 1487, 1341, 1436, 1294, 1390, 1338, 1400, 1325, 1352, 1353, 1288, 1304, 1338, 1355, 1212, 1386, 1426, 1380, 1425, 1287, 1337, 1288, 1348, 1308, 1402, 1370, 1401, 1363, 1312, 1457, 1367, 1320, 1338, 1447, 1371, 1402, 1461, 1382, 1260, 1341, 1309, 1317, 1509, 1403, 1324, 1347, 1351, 1307, 1267, 1312, 1472, 1403, 1327, 1501, 1470, 1438, 1416, 1369, 1355, 1317, 1448, 1423, 1401, 1356, 1400, 1356, 1452, 1435, 1387, 1372, 1390, 1538, 1460, 1474, 1510, 1360, 1424, 1275, 1381, 1453, 1430, 1404, 1350, 1375, 1327, 1312, 1464, 1478, 1536, 1397, 1229, 1337, 1442, 1316, 1455, 1312, 1505, 1440, 1408, 1429, 1280, 1560, 1422, 1363, 1349, 1326, 1400, 1464, 1488, 1352, 1485, 1446, 1540, 1435, 1377, 1287, 1480, 1353, 1359, 1493, 1387, 1314, 1478, 1306, 1462, 1533, 1261, 1488, 1482, 1461, 1452, 1540, 1438, 1423, 1425, 1353, 1489, 1546, 1401, 1459, 1527, 1341, 1516, 1406, 1414, 1442, 1272, 1371, 1435, 1446, 1287, 1496, 1442, 1614, 1305, 1459, 1342, 1478, 1501, 1357, 1428, 1444, 1431, 1425, 1434, 1488, 1508, 1454, 1436, 1485, 1522, 1437, 1396, 1407, 1382, 1444, 1494, 1303, 1552, 1282, 1352, 1412, 1378, 1579, 1543, 1425, 1404, 1380, 1593, 1555, 1532, 1514, 1485, 1504, 1442, 1401, 1453, 1493, 1522, 1417, 1545, 1422, 1540, 1447, 1447, 1575, 1431, 1516, 1542, 1519, 1485, 1526, 1400, 1563, 1471, 1517, 1506, 1514, 1444, 1348, 1588, 1574, 1275, 1331, 1436, 1475, 1570, 1513, 1469, 1573, 1432, 1467, 1513, 1475, 1572, 1430, 1512, 1532, 1487, 1474, 1508, 1410, 1455, 1445, 1544, 1500, 1517, 1496, 1606, 1613, 1526, 1487, 1540, 1511, 1534, 1620, 1409, 1542, 1517, 1493, 1443, 1463, 1391, 1583, 1516, 1700, 1422)
```

--- 

## Lab Questions


### Question 1: Preparation & Visualization

#### Q1a: Create a dataframe that includes the four variables you will need to run a time series model: Passengers, Time, Treatment, and TimeSince. (5 points each)
> Note: Treatment occurs on T=121. TimeSince should be 0 before and on the date of the intervention, and count up from 1 in the periods following.* 

```{r}
# Your code here
```

#### Q1b: Generate summary statistics and plot the Passengers variable over Time. Add a vertical line at T=120 (the day before intervention) to visualize the break point. Briefly describe the trend you see before the intervention. (5 points)
```{r}
# Your code here
```

**Type your answer here, within the asterisks**

--- 

### Question 2: The Naive Model

#### Q2a: Run the standard Interrupted Time Series model using OLS (lm) and print the results in a stargazer table. (10 points)
```{r}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Your code here

```

#### Q2b: Define and interpret the following coefficients, including their significance levels and their practical meaning in the context of our policy intervention: (5 points each)
  > Note: Define what the coefficient represents, its value, significance level, and the meaning of the number in terms of our data context (e.g. a coefficient of 100 would be 100,000 passengers).

**Intercept** ($b_0$): 

  - **Place your answer inside asterisks**
  
**Time** ($b_1$):  

  - **Place your answer inside asterisks**
  
**Treatment** ($b_2$):  

  - **Place your answer inside asterisks**
  
**TimeSince** ($b_3$):

  - **Place your answer inside asterisks**
  

--- 

### Question 3: Diagnostics & Robustness 

#### Q3a: Test for Autocorrelation. Run a Durbin-Watson test (dwtest from lmtest package) and plot the Autocorrelation Function (acf) of your model residuals. Explain the output from both. Based on these two diagnostics, do we satisfy the assumption of independent errors? Is the level of autocorrelation what we would expect to see with time series data? (10 points)
```{r}
# Your code here
```

**Type your answer here, within the asterisks**

#### Q3b: Re-estimate your model using Robust Standard Errors. Use the Newey-West function (from sandwich package) to calculate the robust standard errors. Then, use stargazer to display the Naive Model (from Q2) and the Robust Model side-by-side. (10 points)
```{r}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Your code here

```

#### Q3c: Compare the results. Did the statistical significance of your coefficients change after calculating robust standard errors? What does this imply for the Mayor's policy? (5 points)
> Note: Robust standard errors 'fix' both autocorrelation AND heterskedasticity in our data. If autocorrelation DOES NOT exist and our significance levels change, it must have been due to heteroskedasticity, not autocorrelation.

**Type your answer here, within the asterisks**


--- 

### Question 4: Counterfactual Analysis

#### Q4a: Print the actual number of passengers from the data at T = 150. What does this value represent? (5 points)
> Note: The actual value from the dataset, not the counterfactual

```{r}
# Your code here
```

**Type your answer here, within the asterisks**

#### Q4b: Calculate the counterfactual number of passengers at T=150. What does this value represent? (5 points)
```{r}
# Hint: The predict() function often prints a row number with the result. 
# You can use unname(predict(...)) to remove it for a cleaner output.

# Your code here
```

**Type your answer here, within the asterisks**

#### Q4c: Calculate the counterfactual number of passengers at T=250. Why is the counterfactual from T=150 different from T=250? (5 points)
```{r}
# Hint: The predict() function often prints a row number with the result. 
# You can use unname(predict(...)) to remove it for a cleaner output.

# Your code here
```

**Type your answer here, within the asterisks**

#### Q4d: What is the difference between the actual number of passengers at T=150 and the counterfactual from the same time? In the context of the policy, what does this difference represent? (5 points)
```{r}
# Your code here
```

**Type your answer here, within the asterisks**

---

### Bonus: Control Groups

We have learned that there are threats to the validity of time series analysis. In particular, another event might have occurred at the same time of the intervention and caused the immediate and sustained effect that we observe.

A way to address this issue is to use a control group that is not subject to the intervention. This design makes sure that the effect we observe is the result of the policy intervention.

The mayor proposes to utilize city B as a control group. City B is a neighbor city with very similar characteristics to city A. Yet city B has not changed its bus schedule in the past year.

#### BQ1: Upload the data TS_Groups_Lab.csv from the URL provided below. Estimate a new time series model that includes the control group (Equation 4.1 in your lecture). (5 points)
```{r}
# Loads in Data for Bonus Questions
URL <- "https://raw.githubusercontent.com/DS4PS/pe4ps-textbook/master/labs/DATA/TS_Groups_lab.csv"

# This step simply removes an unnecessary column (X) from the dataframe and renames the passengers column to Passengers
dat.grouped <- read.csv( URL, stringsAsFactors=F) %>% select(passengers, Time, Treatment, TimeSince, Group) %>%  rename(Passengers = passengers)

# Preview the data
rbind(head(dat.grouped, 5), tail(dat.grouped, 5))
```

```{r}
# Reminder: While completing your lab, set type = "text" for your stargazer(), but, remember to set type = "html" when you are ready to knit and submit your lab. You will also need to include in your code chunk header results =' asis' to properly render the table in the knitted HTML file. Your code chunk header should look like this: ```{r, results = 'asis'}

# Your code here

  # Build your ITS w/ Control Group Model
  # Print the results with stargazer
```


#### BQ2: Interpret the Treatment * Group coefficient. Is it statistically significant? (5 points)
**Place your answer inside asterisks**



## Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file.

1. Login to Canvas at http://canvas.asu.edu and navigate to the assignments tab in the course repository.
2. Upload your HTML and RMD files to the appropriate lab submission link.

**Remember to:**

* Include your name as the author in the YAML header.
* Name your files according to the convention: `Lab-##-LastName.Rmd`.
* Show your solution and include your code.
* Format your regression tables using `stargazer`.
* **Bold your answers.**
* Do **NOT** print excessive output (like a full data set).
* Follow appropriate style guidelines (spaces between arguments, etc.).

*Note: Platforms like BlackBoard and Canvas sometimes disallow you from submitting HTML files when there is embedded computer code. If this happens, create a zipped folder with both the RMD and HTML files.*

```
ds4ps
Prog Eval for
Public Service
PE4PS

Labs designed by Federica Fusi & Jesse Lecy
Source code is available on GitHub

Creative Common License:
(CC BY-NC-SA 4.0)

CPP 525 Foundations of Program Eval III
Part of the MS in Evaluation and Analytics
@ Arizona State University

Solutions posted online
are in violation of ASU student policies
and federal copyright laws ©
```
